---
title: "PenicillinX-Project"
format: pdf
toc: true
header-includes:
  - \usepackage[margin=1in]{geometry}
---


# Hutten et al.

```{r OriginalChart, echo=FALSE, include=TRUE, message=FALSE, fig.cap="Cross reactivity matrix re-created from Hutten et al."}
#| fig-width: 10
#| fig-height: 10
#| out-width: "100%"
# Install if missing: BiocManager::install("ComplexHeatmap")
library(ComplexHeatmap)
library(circlize)
# Make sure Gu are cited for the above libraries
# [@gu2014]
# [@gu2022]

data <- read.csv("~/PycharmProjects/PenicillinX-Project/Original.csv", row.names = 1, check.names = FALSE)
mat <- as.matrix(data)

# Makes midline grey
diag(mat) <- -1

# Matches reference dataset
col_fn = colorRamp2(c(-1, 0, 1, 2, 3), c("#808080", "#A9D08F", "#FF0000","#CF734F","#FFFFFF"))


Heatmap(mat, 
        name = "Match",
        col = col_fn,
        
        # Labels
        row_names_side = "left",      # Moves antibiotic labels to the Left
        column_names_side = "top",    # Moves antibiotic labels to the Top
        
        # Grid
        rect_gp = gpar(col = "black", lwd = 0.4), # Adds the black grid lines
        cluster_rows = FALSE,         # Keep your specific list order
        cluster_columns = FALSE,
        
        
        # TEXT 
        column_names_rot = 90,        # Vertical text for top labels
        row_names_gp = gpar(fontsize = 8),
        column_names_gp = gpar(fontsize = 8),
        column_title = "Hutten et al.",
        show_heatmap_legend = FALSE,
        
# Matching the thick lines from reference dataset
        layer_fun = function(j, i, x, y, width, height, fill) {
          n_row = nrow(mat)
          n_col = ncol(mat)
          
          # Vertical lines 
          v_idx = seq(5, n_col - 1, by = 5)
          if(length(v_idx) > 0) {
            grid.segments(
              x0 = unit(v_idx / n_col, "npc"), 
              y0 = unit(0, "npc"),
              x1 = unit(v_idx / n_col, "npc"), 
              y1 = unit(1, "npc"),
              gp = gpar(lwd = 1.4, col = "black")
            )
          }
          
          # Horizontal lines 
          h_idx = seq(5, n_row - 1, by = 5)
          if(length(h_idx) > 0) {
            grid.segments(
              x0 = unit(0, "npc"), 
              y0 = unit(1 - (h_idx / n_row), "npc"),
              x1 = unit(1, "npc"), 
              y1 = unit(1 - (h_idx / n_row), "npc"),
              gp = gpar(lwd = 1.4, col = "black")
            )
          }
        }

)
```

# Direct Matching

```{python directMatchAlgorithm, echo=FALSE}
import pandas as pd
from rdkit import Chem
import numpy as np

abxData = pd.read_csv("antibiotics_data.csv")
ABxList = abxData["Antibiotic"].tolist()

#len(set(abxData["SMILESR1"])) # Can delete, tells unique number of Side chains


mol_series = abxData.set_index("Antibiotic")["SMILESR1"].apply(
    lambda x: Chem.MolFromSmiles(x) if pd.notna(x) else None
)

names = ABxList
matrix_df = pd.DataFrame(0, index=names, columns=names)

for row_name in names:
    row_mol = mol_series[row_name]
    if row_mol is None:
        continue

    # Check both directions: A contains B OR B contains A
    matrix_df.loc[row_name] = mol_series.apply(
        lambda col_mol: 1 if (col_mol and row_mol and (
            row_mol.HasSubstructMatch(col_mol) or
            col_mol.HasSubstructMatch(row_mol)
        )) else 0
    )
np.fill_diagonal(matrix_df.values, -1) # Puts the -1 for the identical antibiotics
matrix_df.to_csv('direct_match.csv', sep=',', index_label='Antibiotic')


```
<!-- Keep adjacent-->
## Direct Match Chart
```{r DirectMatchChart, echo=FALSE, include=TRUE, message=FALSE, fig.cap="Cross reactivity matrix using direct matching algorithm"}
#| fig-width: 10
#| fig-height: 10
#| out-width: "100%"
# Install if missing: BiocManager::install("ComplexHeatmap")
library(ComplexHeatmap)
library(circlize)
# Make sure Gu are cited for the above libraries
# [@gu2014]
# [@gu2022]

data <- read.csv("~/PycharmProjects/PenicillinX-Project/direct_match.csv", row.names = 1, check.names = FALSE)
mat <- as.matrix(data)

# Makes midline grey
diag(mat) <- -1

# Matches reference dataset
col_fn = colorRamp2(c(-1, 0, 1), c("#808080", "#A9D08F", "#FF0000"))


Heatmap(mat, 
        name = "Match",
        col = col_fn,
        
        # Labels
        row_names_side = "left",      # Moves antibiotic labels to the Left
        column_names_side = "top",    # Moves antibiotic labels to the Top
        
        # Grid
        rect_gp = gpar(col = "black", lwd = 0.4), # Adds the black grid lines
        cluster_rows = FALSE,         # Keep your specific list order
        cluster_columns = FALSE,
        
        
        # TEXT 
        column_names_rot = 90,        # Vertical text for top labels
        row_names_gp = gpar(fontsize = 8),
        column_names_gp = gpar(fontsize = 8),
        column_title = "PenicillinX Direct Matching",
        show_heatmap_legend = FALSE,
        
# Matching the thick lines from reference dataset
        layer_fun = function(j, i, x, y, width, height, fill) {
          n_row = nrow(mat)
          n_col = ncol(mat)
          
          # Vertical lines 
          v_idx = seq(5, n_col - 1, by = 5)
          if(length(v_idx) > 0) {
            grid.segments(
              x0 = unit(v_idx / n_col, "npc"), 
              y0 = unit(0, "npc"),
              x1 = unit(v_idx / n_col, "npc"), 
              y1 = unit(1, "npc"),
              gp = gpar(lwd = 1.4, col = "black")
            )
          }
          
          # Horizontal lines 
          h_idx = seq(5, n_row - 1, by = 5)
          if(length(h_idx) > 0) {
            grid.segments(
              x0 = unit(0, "npc"), 
              y0 = unit(1 - (h_idx / n_row), "npc"),
              x1 = unit(1, "npc"), 
              y1 = unit(1 - (h_idx / n_row), "npc"),
              gp = gpar(lwd = 1.4, col = "black")
            )
          }
        }

)
```
<!-- Keep adjacent-->
```{r StatsDirectMatch, echo=FALSE, include=TRUE}
library(reticulate)
# Load and convert to matrix
orig_mat <- as.matrix(read.csv("Original.csv", row.names=1))
algo_mat <- as.matrix(read.csv("direct_match.csv", row.names=1))

# Ensure identical dimensions/names
algo_mat <- algo_mat[rownames(orig_mat), colnames(orig_mat)]

# Calculate Hits
MDhits <- list(
  truePositives = sum(orig_mat == 1 & algo_mat == 1),
  falsePositives = sum(orig_mat == 0 & algo_mat == 1),
  falseNegatives = sum(orig_mat == 1 & algo_mat == 0),
  discrepancyHit = sum(orig_mat == 2 & algo_mat == 1),
  noDataHit     = sum(orig_mat == 3 & algo_mat == 1),
  totalHits    = sum(algo_mat == 1),
  possibleHits = sum(orig_mat ==1)
)


MDhits$hitRate <- round((MDhits$truePositives / MDhits$possibleHits) * 100, 2)

# Reshape for a clean vertical table
directMatchTable <- data.frame(
  Metric = c("Possible Hits (Original)", "Total Hits (Algo)", "True Positives", 
             "False Positives", "False Negatives", "No Data in Original", "Hit Rate (%)"),
  Value  = c(MDhits$possibleHits, MDhits$totalHits, MDhits$truePositives, 
             MDhits$falsePositives, MDhits$falseNegatives, MDhits$noDataHit, MDhits$hitRate)
)



#knitr::kable(directMatchTable,  label = "Direct Matching Results")

#print(MDhits)

```
<!-- Keep adjacent-->

\newpage

Direct Matching calculates: \
A total of `{r} MDhits$totalHits` cross reactions \
out of a possible `{r} MDhits$possibleHits` hits from the original \
This corresponds to `{r} MDhits$truePositives` true positives \
and `{r} MDhits$falsePositives` false positives  \
with `{r} MDhits$falseNegatives` false Negatives \
and `{r} MDhits$noDataHit` for which there is no data in the original \
True positive hitrate is `{r} round(MDhits[[1]] / MDhits[[7]] *100, 2) `%


# Dice Similarity with Morgan Fingerprinting

```{python RDKit_Dice_Threshold, echo=FALSE, include=TRUE}
from rdkit import DataStructs
import numpy as np
from rdkit.Chem import rdFingerprintGenerator
reactiveMatches = 0
similarity_threshold = 0.667
print(f"Morgan Fingerprinting with Dice Similarity (Threshold: {similarity_threshold})")
#print(f"Morgan Fingerprinting with Dice (Threshold: {similarity_threshold})")

# 1. Initialize the modern Fingerprint Generator (Radius 2 = ECFP4)
# fpSize=2048 is the standard default
fp_gen = rdFingerprintGenerator.GetMorganGenerator(radius=2, fpSize=2048)

# 2. Pre-convert SMILES to Mols
mol_series = abxData.set_index("Antibiotic")["SMILESR1"].apply(
    lambda x: Chem.MolFromSmiles(x) if pd.notna(x) else None
)

# 3. Pre-generate Count Fingerprints using the new generator
# GetCountFingerprint(m) creates the count-based vector needed for Dice
fp_series = mol_series.apply(
    lambda m: fp_gen.GetCountFingerprint(m) if m else None
)

names = ABxList
matrix_df = pd.DataFrame(0, index=names, columns=names)

for row_name in names:
    row_fp = fp_series[row_name]
    if row_fp is None:
        continue

    # 4. Calculate Dice Similarity with the threshold
    matrix_df.loc[row_name] = fp_series.apply(
        lambda col_fp: 1 if (col_fp and DataStructs.DiceSimilarity(row_fp, col_fp) >= similarity_threshold) else 0
    )

np.fill_diagonal(matrix_df.values, -1) # Puts the -1 for the identical antibiotics
matrix_df.to_csv('MorganDice.csv', sep=',', index_label='Antibiotic')
```
<!-- Keep Adjacent-->
## Morgan-Dice Chart
```{r MorganDiceChart, echo=FALSE, include=TRUE, message=FALSE, fig.cap="Cross reactivity matrix using Morgan Fingerprinting and Dice Similarity"}
#| fig-width: 10
#| fig-height: 10
#| out-width: "100%"
# Install if missing: BiocManager::install("ComplexHeatmap")
library(ComplexHeatmap)
library(circlize)
# Make sure Gu are cited for the above libraries
# [@gu2014]
# [@gu2022]

data <- read.csv("~/PycharmProjects/PenicillinX-Project/MorganDice.csv", row.names = 1, check.names = FALSE)
mat <- as.matrix(data)

# Makes midline grey
# diag(mat) <- -1 no longer needed

# Matches reference dataset
col_fn = colorRamp2(c(-1, 0, 1), c("#808080", "#A9D08F", "#FF0000"))


Heatmap(mat, 
        name = "Match",
        col = col_fn,
        
        # Labels
        row_names_side = "left",      # Moves antibiotic labels to the Left
        column_names_side = "top",    # Moves antibiotic labels to the Top
        
        # Grid
        rect_gp = gpar(col = "black", lwd = 0.4), # Adds the black grid lines
        cluster_rows = FALSE,         # Keep your specific list order
        cluster_columns = FALSE,
        column_title = "Dice similarity with Morgan Fingerprinting",
        
        # TEXT 
        column_names_rot = 90,        # Vertical text for top labels
        row_names_gp = gpar(fontsize = 8),
        column_names_gp = gpar(fontsize = 8),
        
        show_heatmap_legend = FALSE,
        
# Matching the thick lines from reference dataset
        layer_fun = function(j, i, x, y, width, height, fill) {
          n_row = nrow(mat)
          n_col = ncol(mat)
          
          # Vertical lines 
          v_idx = seq(5, n_col - 1, by = 5)
          if(length(v_idx) > 0) {
            grid.segments(
              x0 = unit(v_idx / n_col, "npc"), 
              y0 = unit(0, "npc"),
              x1 = unit(v_idx / n_col, "npc"), 
              y1 = unit(1, "npc"),
              gp = gpar(lwd = 1.4, col = "black")
            )
          }
          
          # Horizontal lines 
          h_idx = seq(5, n_row - 1, by = 5)
          if(length(h_idx) > 0) {
            grid.segments(
              x0 = unit(0, "npc"), 
              y0 = unit(1 - (h_idx / n_row), "npc"),
              x1 = unit(1, "npc"), 
              y1 = unit(1 - (h_idx / n_row), "npc"),
              gp = gpar(lwd = 1.4, col = "black")
            )
          }
        }

)
```
<!-- Keep adjacent-->
```{r StatsMorganDice, echo=FALSE, include=TRUE}
library(reticulate)
# Load and convert to matrix
orig_mat <- as.matrix(read.csv("Original.csv", row.names=1))
algo_mat <- as.matrix(read.csv("MorganDice.csv", row.names=1))

# Ensure identical dimensions/names
algo_mat <- algo_mat[rownames(orig_mat), colnames(orig_mat)]

# Calculate Hits
MDhits <- list(
  truePositives = sum(orig_mat == 1 & algo_mat == 1),
  falsePositives = sum(orig_mat == 0 & algo_mat == 1),
  falseNegatives = sum(orig_mat == 1 & algo_mat == 0),
  discrepancyHit = sum(orig_mat == 2 & algo_mat == 1),
  noDataHit     = sum(orig_mat == 3 & algo_mat == 1),
  totalHits    = sum(algo_mat == 1),
  possibleHits = sum(orig_mat ==1)
)

threshold_value <- py$similarity_threshold

MDhits$hitRate <- round((MDhits$truePositives / MDhits$possibleHits) * 100, 2)

# Reshape for a clean vertical table
morganDiceTable <- data.frame(
  Metric = c("Possible Hits (Original)", "Total Hits (Algo)", "True Positives", 
             "False Positives", "False Negatives", "No Data in Original", "Hit Rate (%)"),
  Value  = c(MDhits$possibleHits, MDhits$totalHits, MDhits$truePositives, 
             MDhits$falsePositives, MDhits$falseNegatives, MDhits$noDataHit, MDhits$hitRate)
)



#knitr::kable(morganDiceTable,  label = "PharmacoDice Results")


```
<!-- Keep adjacent-->

\newpage

Dice similarity with Morgan Fingerprinting at a similarity threshold of `{r} threshold_value` calculates: \
A total of `{r} MDhits$totalHits` cross reactions \
out of a possible `{r} MDhits$possibleHits` hits from the original \
This corresponds to `{r} MDhits$truePositives` true positives \
and `{r} MDhits$falsePositives` false positives  \
with `{r} MDhits$falseNegatives` false Negatives \
and `{r} MDhits$noDataHit` for which there is no data in the original \
True positive hitrate is `{r} round(MDhits[[1]] / MDhits[[7]] *100, 2) `%






# Tanimoto similarity

```{python MorganTanimoto, echo=FALSE, include=TRUE}
from rdkit import DataStructs
import numpy as np
from rdkit.Chem import rdFingerprintGenerator
reactiveMatches = 0
similarity_threshold = 0.5
print(f"Morgan Fingerprinting with Tanimoto Similarity (Threshold: {similarity_threshold})")
#print(f"Morgan Fingerprinting with Dice (Threshold: {similarity_threshold})")

# 1. Initialize the modern Fingerprint Generator (Radius 2 = ECFP4)
# fpSize=2048 is the standard default
fp_gen = rdFingerprintGenerator.GetMorganGenerator(radius=2, fpSize=2048)

# 2. Pre-convert SMILES to Mols
mol_series = abxData.set_index("Antibiotic")["SMILESR1"].apply(
    lambda x: Chem.MolFromSmiles(x) if pd.notna(x) else None
)

# 3. Pre-generate Count Fingerprints using the new generator
# GetCountFingerprint(m) creates the count-based vector needed for Dice
fp_series = mol_series.apply(
    lambda m: fp_gen.GetCountFingerprint(m) if m else None
)

names = ABxList
matrix_df = pd.DataFrame(0, index=names, columns=names)

for row_name in names:
    row_fp = fp_series[row_name]
    if row_fp is None:
        continue

    # 4. Calculate Dice Similarity with the threshold
    matrix_df.loc[row_name] = fp_series.apply(
        lambda col_fp: 1 if (col_fp and DataStructs.TanimotoSimilarity(row_fp, col_fp) >= similarity_threshold) else 0
    )

np.fill_diagonal(matrix_df.values, -1) # Puts the -1 for the identical antibiotics
matrix_df.to_csv('MorganTanimoto.csv', sep=',', index_label='Antibiotic')
```
<!-- Keep Adjacent-->
## Morgan-Tanimoto Chart
```{r MorganTanimotoChart, echo=FALSE, include=TRUE, message=FALSE, fig.cap="Cross reactivity matrix using Morgan Fingerprinting and Dice Similarity"}
#| fig-width: 10
#| fig-height: 10
#| out-width: "100%"
# Install if missing: BiocManager::install("ComplexHeatmap")
library(ComplexHeatmap)
library(circlize)
# Make sure Gu are cited for the above libraries
# [@gu2014]
# [@gu2022]

data <- read.csv("~/PycharmProjects/PenicillinX-Project/MorganTanimoto.csv", row.names = 1, check.names = FALSE)
mat <- as.matrix(data)

# Makes midline grey
# diag(mat) <- -1 no longer needed

# Matches reference dataset
col_fn = colorRamp2(c(-1, 0, 1), c("#808080", "#A9D08F", "#FF0000"))


Heatmap(mat, 
        name = "Match",
        col = col_fn,
        
        # Labels
        row_names_side = "left",      # Moves antibiotic labels to the Left
        column_names_side = "top",    # Moves antibiotic labels to the Top
        
        # Grid
        rect_gp = gpar(col = "black", lwd = 0.4), # Adds the black grid lines
        cluster_rows = FALSE,         # Keep your specific list order
        cluster_columns = FALSE,
        column_title = "Tanimoto similarity with Morgan Fingerprinting",
        
        # TEXT 
        column_names_rot = 90,        # Vertical text for top labels
        row_names_gp = gpar(fontsize = 8),
        column_names_gp = gpar(fontsize = 8),
        
        show_heatmap_legend = FALSE,
        
# Matching the thick lines from reference dataset
        layer_fun = function(j, i, x, y, width, height, fill) {
          n_row = nrow(mat)
          n_col = ncol(mat)
          
          # Vertical lines 
          v_idx = seq(5, n_col - 1, by = 5)
          if(length(v_idx) > 0) {
            grid.segments(
              x0 = unit(v_idx / n_col, "npc"), 
              y0 = unit(0, "npc"),
              x1 = unit(v_idx / n_col, "npc"), 
              y1 = unit(1, "npc"),
              gp = gpar(lwd = 1.4, col = "black")
            )
          }
          
          # Horizontal lines 
          h_idx = seq(5, n_row - 1, by = 5)
          if(length(h_idx) > 0) {
            grid.segments(
              x0 = unit(0, "npc"), 
              y0 = unit(1 - (h_idx / n_row), "npc"),
              x1 = unit(1, "npc"), 
              y1 = unit(1 - (h_idx / n_row), "npc"),
              gp = gpar(lwd = 1.4, col = "black")
            )
          }
        }

)
```
<!-- Keep adjacent-->
```{r StatsMorganTanimoto, echo=FALSE, include=TRUE}
#| tbl-cap: "Results"
library(gt)

# Load and convert to matrix
orig_mat <- as.matrix(read.csv("Original.csv", row.names=1))
algo_mat <- as.matrix(read.csv("MorganTanimoto.csv", row.names=1))

# Ensure identical dimensions/names
algo_mat <- algo_mat[rownames(orig_mat), colnames(orig_mat)]

# Calculate Hits
MDhits <- list(
  possibleHits = sum(orig_mat ==1),
  totalHits    = sum(algo_mat == 1),
  truePositives = sum(orig_mat == 1 & algo_mat == 1),
  falsePositives = sum(orig_mat == 0 & algo_mat == 1),
  falseNegatives = sum(orig_mat == 1 & algo_mat == 0),
  discrepancyHit = sum(orig_mat == 2 & algo_mat == 1),
  noDataHit     = sum(orig_mat == 3 & algo_mat == 1)
)


threshold_value <- py$similarity_threshold

MDhits$hitRate <- round((MDhits$truePositives / MDhits$possibleHits) * 100, 2)

# Reshape for a clean vertical table
morganTanimotoTable <- data.frame(
  Metric = c("Possible Hits (Original)", "Total Hits (Algo)", "True Positives", 
             "False Positives", "False Negatives", "No Data in Original", "Hit Rate (%)"),
  Value  = c(MDhits$possibleHits, MDhits$totalHits, MDhits$truePositives, 
             MDhits$falsePositives, MDhits$falseNegatives, MDhits$noDataHit, MDhits$hitRate)
)



#knitr::kable(morganTanimotoTable,  caption = "Cross-Reactivity Analysis Results")


#print(MDhits)





```
<!-- Keep adjacent-->

\newpage
<!--
Dice similarity with Tanimoto Fingerprinting at a similarity threshold of `{r} MDhits$threshold_value` calculates: \
A total of `{r} MDhits$totalHits` cross reactions \
out of a possible `{r} MDhits$possibleHits` hits from the original \
This corresponds to `{r} MDhits$truePositives` true positives \
and `{r} MDhits$falsePositives` false positives  \
with `{r} MDhits$falseNegatives` false Negatives \
and `{r} MDhits$noDataHit` for which there is no data in the original \
True positive hitrate is `{r} round(MDhits[[3]] / MDhits[[1]] *100, 2) `%


-->


# Pharmacophore fingerprinting

```{python  PharmacophoreDice, echo=FALSE, include=TRUE}
from rdkit import Chem
from rdkit import DataStructs
from rdkit.Chem.Pharm2D import Gobbi_Pharm2D, Generate
import pandas as pd
import numpy as np

# Parameters
similarity_threshold = 0.667
print(f"Pharmacophore Fingerprinting with Dice (Threshold: {similarity_threshold})")

# 1. Access the pre-configured Gobbi factory
factory = Gobbi_Pharm2D.factory

# 2. Pre-convert SMILES to Mols
mol_series = abxData.set_index("Antibiotic")["SMILESR1"].apply(
    lambda x: Chem.MolFromSmiles(x) if pd.notna(x) else None
)

# 3. Pre-generate Pharmacophore Fingerprints
# Using Gen2DFingerprint as verified in your working example
fp_series = mol_series.apply(
    lambda m: Generate.Gen2DFingerprint(m, factory) if m else None
)

names = ABxList
matrix_df = pd.DataFrame(0, index=names, columns=names)

# 4. Compute the similarity matrix
for row_name in names:
    row_fp = fp_series[row_name]
    if row_fp is None:
        continue

    # DiceSimilarity is calculated between the pharmacophore fingerprints
    matrix_df.loc[row_name] = fp_series.apply(
        lambda col_fp: 1 if (col_fp and DataStructs.DiceSimilarity(row_fp, col_fp) >= similarity_threshold) else 0
    )

# 5. Mirror the R identity diagonal
np.fill_diagonal(matrix_df.values, -1)

# 6. Export to CSV
matrix_df.to_csv('PharmacoDice.csv', sep=',', index_label='Antibiotic')
    
```
<!--Keep adjacent-->
## Pharmacophore-Dice Chart
```{r PharmacophoreChart, echo=FALSE, include=TRUE, message=FALSE, fig.cap="Cross reactivity matrix using Pharmacophore and Dice Similarity"}
#| fig-width: 10
#| fig-height: 10
#| out-width: "100%"
# Install if missing: BiocManager::install("ComplexHeatmap")
library(ComplexHeatmap)
library(circlize)
# Make sure Gu are cited for the above libraries
# [@gu2014]
# [@gu2022]

data <- read.csv("~/PycharmProjects/PenicillinX-Project/PharmacoDice.csv", row.names = 1, check.names = FALSE)
mat <- as.matrix(data)

# Makes midline grey
# diag(mat) <- -1 no longer needed

# Matches reference dataset
col_fn = colorRamp2(c(-1, 0, 1), c("#808080", "#A9D08F", "#FF0000"))


Heatmap(mat, 
        name = "Match",
        col = col_fn,
        
        # Labels
        row_names_side = "left",      # Moves antibiotic labels to the Left
        column_names_side = "top",    # Moves antibiotic labels to the Top
        
        # Grid
        rect_gp = gpar(col = "black", lwd = 0.4), # Adds the black grid lines
        cluster_rows = FALSE,         # Keep your specific list order
        cluster_columns = FALSE,
        column_title = "Pharmacophore similarity with Dice Fingerprinting",
        
        # TEXT 
        column_names_rot = 90,        # Vertical text for top labels
        row_names_gp = gpar(fontsize = 8),
        column_names_gp = gpar(fontsize = 8),
        
        show_heatmap_legend = FALSE,
        
# Matching the thick lines from reference dataset
        layer_fun = function(j, i, x, y, width, height, fill) {
          n_row = nrow(mat)
          n_col = ncol(mat)
          
          # Vertical lines 
          v_idx = seq(5, n_col - 1, by = 5)
          if(length(v_idx) > 0) {
            grid.segments(
              x0 = unit(v_idx / n_col, "npc"), 
              y0 = unit(0, "npc"),
              x1 = unit(v_idx / n_col, "npc"), 
              y1 = unit(1, "npc"),
              gp = gpar(lwd = 1.4, col = "black")
            )
          }
          
          # Horizontal lines 
          h_idx = seq(5, n_row - 1, by = 5)
          if(length(h_idx) > 0) {
            grid.segments(
              x0 = unit(0, "npc"), 
              y0 = unit(1 - (h_idx / n_row), "npc"),
              x1 = unit(1, "npc"), 
              y1 = unit(1 - (h_idx / n_row), "npc"),
              gp = gpar(lwd = 1.4, col = "black")
            )
          }
        }

)
```
<!-- Keep adjacent-->
```{r StatsPharmacoDice, echo=FALSE, include=TRUE}
#| tbl-cap: "Results"
library(gt)

# Load and convert to matrix
orig_mat <- as.matrix(read.csv("Original.csv", row.names=1))
algo_mat <- as.matrix(read.csv("PharmacoDice.csv", row.names=1))

# Ensure identical dimensions/names
algo_mat <- algo_mat[rownames(orig_mat), colnames(orig_mat)]

# Calculate Hits
MDhits <- list(
  possibleHits = sum(orig_mat ==1),
  totalHits    = sum(algo_mat == 1),
  truePositives = sum(orig_mat == 1 & algo_mat == 1),
  falsePositives = sum(orig_mat == 0 & algo_mat == 1),
  falseNegatives = sum(orig_mat == 1 & algo_mat == 0),
  discrepancyHit = sum(orig_mat == 2 & algo_mat == 1),
  noDataHit     = sum(orig_mat == 3 & algo_mat == 1)
)


threshold_value <- py$similarity_threshold

MDhits$hitRate <- round((MDhits$truePositives / MDhits$possibleHits) * 100, 2)

# Reshape for a clean vertical table
pharmacoDiceTable <- data.frame(
  Metric = c("Possible Hits (Original)", "Total Hits (Algo)", "True Positives", 
             "False Positives", "False Negatives", "No Data in Original", "Hit Rate (%)"),
  Value  = c(MDhits$possibleHits, MDhits$totalHits, MDhits$truePositives, 
             MDhits$falsePositives, MDhits$falseNegatives, MDhits$noDataHit, MDhits$hitRate)
)



#knitr::kable(pharmacoDiceTable,  label = "PharmacoDice Results")


#print(MDhits)





```


# Results

```{r}
knitr::kable(directMatchTable,  label = "Direct Match Results")
knitr::kable(morganDiceTable,  label = "Morgan-Dice Results")
knitr::kable(morganTanimotoTable,  label = "Morgan-Tanimoto Results")
knitr::kable(pharmacoDiceTable,  label = "PharmacoDice Results")
```

