---
title: "PenicillinX-Project"
format: pdf
toc: true
header-includes:
  - \usepackage[margin=1in]{geometry}
---




```{r echo=FALSE, include=TRUE, message=FALSE, fig.cap="Cross reactivity matrix re-created from Hutten et al."}
#| label: OriginalChart
library(ComplexHeatmap)
library(circlize)
# Make sure Gu are cited for the above libraries
# [@gu2014]
# [@gu2022]
huttenMat <- as.matrix(read.csv("~/PycharmProjects/PenicillinX-Project/Original.csv", row.names = 1, check.names = FALSE))
# Function to calculate the stats
calculateStats <- function(csvPath, originalMat) {
  algoMat <- as.matrix(read.csv(csvPath, row.names = 1, check.names = FALSE))
  algoMat <- algoMat[rownames(originalMat), colnames(originalMat)]
  
  list(
    possibleHits   = sum(originalMat == 1),
    totalHits      = sum(algoMat == 1),
    truePositives  = sum(originalMat == 1 & algoMat == 1),
    trueNegatives  = sum(originalMat == 0 & algoMat == 0),
    falsePositives = sum(originalMat == 0 & algoMat == 1),
    falseNegatives = sum(originalMat == 1 & algoMat == 0),
    noDataHit      = sum(originalMat == 3 & algoMat == 1),
 #   hitRate        = round((sum(originalMat == 1 & algoMat == 1) / sum(originalMat == 1)),  2),
    sensitivity    = round(sum(originalMat == 1 & algoMat == 1) / (sum(originalMat == 1 & algoMat == 1) + sum(originalMat == 1 & algoMat == 0)),2),
    specificity    = round(sum(originalMat == 0 & algoMat == 0) / (sum(originalMat == 0 & algoMat == 0) + sum(originalMat == 0 & algoMat == 1)),3)
  )
}


# Function to generate Heatmap
plotAbxHeatmap <- function(csvPath, chartTitle, colMapping) {
  localMat <- as.matrix(read.csv(csvPath, row.names = 1, check.names = FALSE))
  
  Heatmap(localMat, 
          name = "Match", 
          col = colMapping,  # Now uses the mapping passed into the function
          column_title = chartTitle,
          row_names_side = "left", 
          column_names_side = "top",
          cluster_rows = FALSE, 
          cluster_columns = FALSE,
          rect_gp = gpar(col = "black", lwd = 0.4),
          column_names_rot = 90, 
          row_names_gp = gpar(fontsize = 8),
          column_names_gp = gpar(fontsize = 8), 
          show_heatmap_legend = FALSE,
          
          layer_fun = function(j, i, x, y, width, height, fill) {
            nRow = nrow(localMat)
            nCol = ncol(localMat)
            vIdx = seq(5, nCol - 1, by = 5)
            if(length(vIdx) > 0) {
              grid.segments(x0 = unit(vIdx / nCol, "npc"), y0 = unit(0, "npc"),
                            x1 = unit(vIdx / nCol, "npc"), y1 = unit(1, "npc"),
                            gp = gpar(lwd = 1.4, col = "black"))
            }
            hIdx = seq(5, nRow - 1, by = 5)
            if(length(hIdx) > 0) {
              grid.segments(x0 = unit(0, "npc"), y0 = unit(1 - (hIdx / nRow), "npc"),
                            x1 = unit(1, "npc"), y1 = unit(1 - (hIdx / nRow), "npc"),
                            gp = gpar(lwd = 1.4, col = "black"))
            }
          })
}
```

@fig-huttenChart is a reconstruction of the matrix detailed in Hutten et al. Below them are the various molecular matching algorithms and their respective matrices, the performance of each algorithm is displayed in a table at the end.


<!--## algoblock -->

```{python algoBlock, include=TRUE, echo=FALSE}
import pandas as pd
import numpy as np
from rdkit import Chem, DataStructs
from rdkit.Chem import rdFingerprintGenerator, MACCSkeys, rdFMCS
from rdkit.Chem.Pharm2D import Gobbi_Pharm2D, Generate

###### Adjust thresholds here ######
thresholdMD = 0.667 # 100% Specificity at 0.667
thresholdMT = 0.667 # 100% Specificity at 0.667
thresholdPhD = 0.667 # 100% Specificity at 0.667
thresholdMAC = 0.667 # 100% Specificity at 0.667
thresholdMCS = 0.95 #  # 100% Specificity at TBD
###### -------------------- #######

# Pull initial antibiotic data
abxData = pd.read_csv("antibiotics_data.csv")
names = abxData["Antibiotic"].tolist()
molSeries = abxData.set_index("Antibiotic")["SMILESR1"].apply(
    lambda x: Chem.MolFromSmiles(x) if pd.notna(x) else None
)

def save_matrix(df, fileName):
    np.fill_diagonal(df.values, -1)
    df.to_csv(fileName, index_label='Antibiotic')

# 1. Direct Match
directMatchDf = pd.DataFrame(0, index=names, columns=names)
for row in names:
    m1 = molSeries[row]
    if m1:
        directMatchDf.loc[row] = molSeries.apply(
            lambda m2: 1 if (m2 and (m1.HasSubstructMatch(m2) or m2.HasSubstructMatch(m1))) else 0
        )
save_matrix(directMatchDf, 'directMatch.csv')

# 2. Morgan Fingerprints (Dice & Tanimoto)
fpGen = rdFingerprintGenerator.GetMorganGenerator(radius=2, fpSize=2048)
fpSeries = molSeries.apply(lambda m: fpGen.GetCountFingerprint(m) if m else None)

for mode, threshold, fileName in [('Dice', thresholdMD, 'morganDice.csv'), ('Tanimoto', thresholdMT, 'morganTanimoto.csv')]:
    df = pd.DataFrame(0, index=names, columns=names)
    simFunc = DataStructs.DiceSimilarity if mode == 'Dice' else DataStructs.TanimotoSimilarity
    for row in names:
        f1 = fpSeries[row]
        if f1:
            df.loc[row] = fpSeries.apply(lambda f2: 1 if (f2 and simFunc(f1, f2) >= threshold) else 0)
    save_matrix(df, fileName)

# 3. Pharmacophore
pharmFpSeries = molSeries.apply(lambda m: Generate.Gen2DFingerprint(m, Gobbi_Pharm2D.factory) if m else None)
pharmDf = pd.DataFrame(0, index=names, columns=names)
for row in names:
    f1 = pharmFpSeries[row]
    if f1:
        pharmDf.loc[row] = pharmFpSeries.apply(
            lambda f2: 1 if (f2 and DataStructs.DiceSimilarity(f1, f2) >= thresholdPhD) else 0
        )
save_matrix(pharmDf, 'pharmacoDice.csv')

# 4. MACCS Keys 
maccsFpSeries = molSeries.apply(lambda m: MACCSkeys.GenMACCSKeys(m) if m else None)
maccsDf = pd.DataFrame(0, index=names, columns=names)
for row in names:
    f1 = maccsFpSeries[row]
    if f1:
        maccsDf.loc[row] = maccsFpSeries.apply(
            lambda f2: 1 if (f2 and DataStructs.TanimotoSimilarity(f1, f2) >= thresholdMAC) else 0
        )
save_matrix(maccsDf, 'maccsTanimoto.csv')

# 5. Maximum Common Substructure (MCS)
# use atom counts to determine if a significant core is shared
mcsDf = pd.DataFrame(0, index=names, columns=names)
mcsThreshold = thresholdMCS

for i, rowName in enumerate(names):
    m1 = molSeries[rowName]
    if not m1: continue
    for j, colName in enumerate(names):
        if i >= j: continue # save compute by only comparing top right
        m2 = molSeries[colName]
        if m2:
            res = rdFMCS.FindMCS([m1, m2])
            # Similarity in this case is number of shared atoms / atoms in the smaller molecule
            overlap = res.numAtoms / min(m1.GetNumAtoms(), m2.GetNumAtoms())
            if overlap >= mcsThreshold:
                mcsDf.iloc[i, j] = 1
                mcsDf.iloc[j, i] = 1

save_matrix(mcsDf, 'mcsOverlap.csv')

```

# Hutten et al.

```{r echo=FALSE, include=TRUE, message=FALSE, fig.cap="Cross reactivity matrix using direct matching algorithm"}
#| label: fig-huttenChart
#| fig-width: 10
#| fig-height: 10
plotAbxHeatmap("Original.csv", "Hutten et al.", colorRamp2(c(-1, 0, 1, 2, 3), c("#808080", "#A9D08F","#FF0000","#CF734F","#FFFFFF")))
```


# Direct Matching (PenicillinX original)

```{r DirectMatchChart, echo=FALSE, include=TRUE, message=FALSE, fig.cap="Cross reactivity matrix using direct matching algorithm"}
#| fig-width: 10
#| fig-height: 10
plotAbxHeatmap("directMatch.csv", "Direct Matching", colorRamp2(c(-1, 0, 1), c("#808080", "#A9D08F", "#FF0000")))
directMatchStats <- calculateStats("directMatch.csv", huttenMat) # 'mat' is from Hutten et al.
```


# Dice Similarity with Morgan Fingerprinting

```{r morganDiceChart, include=TRUE, echo=FALSE, fig.cap="Cross reactivity matrix using Morgan fingerprinting with Dice Similarity "}
#| fig-width: 10
#| fig-height: 10
plotAbxHeatmap("morganDice.csv", "Morgan-Dice Similarity", colorRamp2(c(-1, 0, 1), c("#808080", "#A9D08F", "#FF0000")))
morganDiceStats <- calculateStats("morganDice.csv", huttenMat)
```


# Tanimoto Similarity with Morgan Fingerprinting

```{r morganTanimotoChart, include=TRUE, echo=FALSE, fig.cap="Cross reactivity matrix using Morgan fingerprinting with Tanimoto Similarity "}
#| fig-width: 10
#| fig-height: 10
plotAbxHeatmap("morganTanimoto.csv", "Morgan-Tanimoto Similarity", colorRamp2(c(-1, 0, 1), c("#808080", "#A9D08F", "#FF0000")))
morganTanimotoStats <- calculateStats("morganTanimoto.csv", huttenMat)
```


# Pharmacocore with Morgan Fingerprinting

```{r PharmacoDice, include=TRUE, echo=FALSE, fig.cap="Cross reactivity matrix using Morgan fingerprinting with Pharmacocores"}
#| fig-width: 10
#| fig-height: 10
plotAbxHeatmap("pharmacoDice.csv", "Pharmacocore Dice Similarity", colorRamp2(c(-1, 0, 1), c("#808080", "#A9D08F", "#FF0000")))
pharmacoDiceStats <- calculateStats("pharmacoDice.csv", huttenMat)
```

# MACCS Tanimoto

```{r MACCSTanimoto, include=TRUE, echo=FALSE, fig.cap="Cross reactivity matrix using MACCS keys with Tanimoto Similarity "}
#| fig-width: 10
#| fig-height: 10
plotAbxHeatmap("maccsTanimoto.csv", "MACCS Tanimoto", colorRamp2(c(-1, 0, 1), c("#808080", "#A9D08F", "#FF0000")))
maccsTanimotoStats <- calculateStats("maccsTanimoto.csv", huttenMat)
```


# MCS Overlap

```{r MCS, include=TRUE, echo=FALSE, fig.cap="Cross reactivity matrix using Maximum Common Substructure"}
#| fig-width: 10
#| fig-height: 10
plotAbxHeatmap("mcsOverlap.csv", "Maximum Common Substructure Overlap", colorRamp2(c(-1, 0, 1), c("#808080", "#A9D08F", "#FF0000")))
mcsStats <- calculateStats("mcsOverlap.csv", huttenMat)
```

\newpage

```{python, echo=FALSE, include=TRUE}
#| label: performance-table
#| tbl-cap: "Compared Performance of Each Algorithm"

from IPython.display import Markdown
from tabulate import tabulate

# Access R lists using the 'r' prefix
# These behave like dictionaries in Python
dm = r.directMatchStats
md = r.morganDiceStats
mt = r.morganTanimotoStats
pd = r.pharmacoDiceStats
mcs = r.mcsStats
mct = r.maccsTanimotoStats

# Define the rows of the table
table = [
    ["Set Threshold", ['N/A'], [thresholdMD], ['thresholdMT'],  ['thresholdPhD'], ['thresholdMCS'],  ['thresholdMAC']],   
    ["Possible Hits", dm['possibleHits'], md['possibleHits'], mt['possibleHits'], pd['possibleHits'], mcs['possibleHits'], mct["possibleHits"]],
    ["Total Hits",    dm['totalHits'],    md['totalHits'],    mt['totalHits'],    pd['totalHits'], mcs['totalHits'], mct["totalHits"]],
    ["True +ve",dm['truePositives'],md['truePositives'],mt['truePositives'], pd['truePositives'], mcs['truePositives'], mct["truePositives"]],
    ["False +ve",dm['falsePositives'],md['falsePositives'],mt['falsePositives'],pd['falsePositives'], mcs['falsePositives'], mct["falsePositives"]],
    ["False -ve",dm['falseNegatives'],md['falseNegatives'],mt['falseNegatives'],pd['falseNegatives'], mcs['falseNegatives'], mct["falseNegatives"]],
    ["no Data", dm['noDataHit'], md['noDataHit'], mt['noDataHit'], pd['noDataHit'], mcs['noDataHit'], mct["noDataHit"]],
  #  ["Hit Rate", dm['hitRate'], md['hitRate'], mt['hitRate'], pd['hitRate'], mcs['hitRate'], mct["hitRate"]],
    ["Sensitivity", dm['sensitivity'], md['sensitivity'], mt['sensitivity'], pd['sensitivity'], mcs['sensitivity'], mct['sensitivity']],
    ["Specificity", dm['specificity'], md['specificity'], mt['specificity'], pd['specificity'], mcs['specificity'], mct['specificity']]
    ]


# Create the headers
headers = ["Metric", "Direct Match", "Morgan Dice", "Morgan Tanimoto", "Pharmaco Dice", "MCS", "MACCS Tanimoto"]

# Render the table
Markdown(tabulate(table, headers=headers, tablefmt="pipe"))
```










