---
title: "PenicillinX-Project"
format: pdf
toc: true
header-includes:
  - \usepackage[margin=1in]{geometry}
---



# Hutten et al.

```{r echo=FALSE, include=TRUE, message=FALSE, fig.cap="Cross reactivity matrix re-created from Hutten et al."}
#| label: OriginalChart
library(ComplexHeatmap)
library(circlize)
# Make sure Gu are cited for the above libraries
# [@gu2014]
# [@gu2022]
huttenMat <- as.matrix(read.csv("~/PycharmProjects/PenicillinX-Project/Original.csv", row.names = 1, check.names = FALSE))
# Function to calculate the stats
calculateStats <- function(csvPath, originalMat) {
  algoMat <- as.matrix(read.csv(csvPath, row.names = 1, check.names = FALSE))
  algoMat <- algoMat[rownames(originalMat), colnames(originalMat)]
  
  list(
    possibleHits   = sum(originalMat == 1) /2,
    totalHits      = sum(algoMat == 1) /2,
    truePositives  = sum(originalMat == 1 & algoMat == 1) /2,
    trueNegatives  = sum(originalMat == 0 & algoMat == 0) /2,
    falsePositives = sum(originalMat == 0 & algoMat == 1) /2,
    falseNegatives = sum(originalMat == 1 & algoMat == 0) /2,
    noDataHit      = sum(originalMat == 3 & algoMat == 1) /2,
 #   hitRate        = round((sum(originalMat == 1 & algoMat == 1) / sum(originalMat == 1)),  2),
    sensitivity    = round(sum(originalMat == 1 & algoMat == 1) / (sum(originalMat == 1 & algoMat == 1) + sum(originalMat == 1 & algoMat == 0)),2),
    specificity    = round(sum(originalMat == 0 & algoMat == 0) / (sum(originalMat == 0 & algoMat == 0) + sum(originalMat == 0 & algoMat == 1)),3)
  )
}


# Function to generate Heatmap
plotAbxHeatmap <- function(csvPath, chartTitle, colMapping) {
  localMat <- as.matrix(read.csv(csvPath, row.names = 1, check.names = FALSE))
  
  Heatmap(localMat, 
          name = "Match", 
          col = colMapping,  # Now uses the mapping passed into the function
          column_title = chartTitle,
          row_names_side = "left", 
          column_names_side = "top",
          cluster_rows = FALSE, 
          cluster_columns = FALSE,
          rect_gp = gpar(col = "black", lwd = 0.4),
          column_names_rot = 90, 
          row_names_gp = gpar(fontsize = 8),
          column_names_gp = gpar(fontsize = 8), 
          show_heatmap_legend = FALSE,
          
          layer_fun = function(j, i, x, y, width, height, fill) {
            nRow = nrow(localMat)
            nCol = ncol(localMat)
            vIdx = seq(5, nCol - 1, by = 5)
            if(length(vIdx) > 0) {
              grid.segments(x0 = unit(vIdx / nCol, "npc"), y0 = unit(0, "npc"),
                            x1 = unit(vIdx / nCol, "npc"), y1 = unit(1, "npc"),
                            gp = gpar(lwd = 1.4, col = "black"))
            }
            hIdx = seq(5, nRow - 1, by = 5)
            if(length(hIdx) > 0) {
              grid.segments(x0 = unit(0, "npc"), y0 = unit(1 - (hIdx / nRow), "npc"),
                            x1 = unit(1, "npc"), y1 = unit(1 - (hIdx / nRow), "npc"),
                            gp = gpar(lwd = 1.4, col = "black"))
            }
          })
}
```

@fig-huttenChart is a reconstruction of the matrix detailed in Hutten et al. Below them are the various molecular matching algorithms and their respective matrices, the performance of each algorithm, as well as their combined performance matching against Hutten et al. is displayed in a table at the end of the document.




```{python algoBlock, include=TRUE, echo=FALSE}
import pandas as pd
import numpy as np
from rdkit import Chem, DataStructs
from rdkit.Chem import rdFingerprintGenerator, MACCSkeys, rdFMCS
from rdkit.Chem.Pharm2D import Gobbi_Pharm2D, Generate

###### Adjust thresholds here ######
thresholdMD = 0.667 # 100% Specificity at 0.667
thresholdMT = 0.667 # 100% Specificity at 0.667
thresholdPhD = 0.667 # 100% Specificity at 0.667
thresholdMAC = 0.667 # 100% Specificity at 0.667
thresholdMCS = 0.95 #  # 100% Specificity at TBD
###### -------------------- #######

# Pull initial antibiotic data
abxData = pd.read_csv("antibiotics_data.csv")
names = abxData["Antibiotic"].tolist()
molSeries = abxData.set_index("Antibiotic")["SMILESR1"].apply(
    lambda x: Chem.MolFromSmiles(x) if pd.notna(x) else None
)

def save_matrix(df, fileName):
    np.fill_diagonal(df.values, -1)
    df.to_csv(fileName, index_label='Antibiotic')

# 1. Direct Match
directMatchDf = pd.DataFrame(0, index=names, columns=names)
for row in names:
    m1 = molSeries[row]
    if m1:
        directMatchDf.loc[row] = molSeries.apply(
            lambda m2: 1 if (m2 and (m1.HasSubstructMatch(m2) or m2.HasSubstructMatch(m1))) else 0
        )
save_matrix(directMatchDf, 'directMatch.csv')

# 2. Morgan Fingerprints (Dice & Tanimoto)
fpGen = rdFingerprintGenerator.GetMorganGenerator(radius=2, fpSize=2048)
fpSeries = molSeries.apply(lambda m: fpGen.GetCountFingerprint(m) if m else None)

for mode, threshold, fileName in [('Dice', thresholdMD, 'morganDice.csv'), ('Tanimoto', thresholdMT, 'morganTanimoto.csv')]:
    df = pd.DataFrame(0, index=names, columns=names)
    simFunc = DataStructs.DiceSimilarity if mode == 'Dice' else DataStructs.TanimotoSimilarity
    for row in names:
        f1 = fpSeries[row]
        if f1:
            df.loc[row] = fpSeries.apply(lambda f2: 1 if (f2 and simFunc(f1, f2) >= threshold) else 0)
    save_matrix(df, fileName)

# 3. Pharmacophore
pharmFpSeries = molSeries.apply(lambda m: Generate.Gen2DFingerprint(m, Gobbi_Pharm2D.factory) if m else None)
pharmDf = pd.DataFrame(0, index=names, columns=names)
for row in names:
    f1 = pharmFpSeries[row]
    if f1:
        pharmDf.loc[row] = pharmFpSeries.apply(
            lambda f2: 1 if (f2 and DataStructs.DiceSimilarity(f1, f2) >= thresholdPhD) else 0
        )
save_matrix(pharmDf, 'pharmacoDice.csv')

# 4. MACCS Keys 
maccsFpSeries = molSeries.apply(lambda m: MACCSkeys.GenMACCSKeys(m) if m else None)
maccsDf = pd.DataFrame(0, index=names, columns=names)
for row in names:
    f1 = maccsFpSeries[row]
    if f1:
        maccsDf.loc[row] = maccsFpSeries.apply(
            lambda f2: 1 if (f2 and DataStructs.TanimotoSimilarity(f1, f2) >= thresholdMAC) else 0
        )
save_matrix(maccsDf, 'maccsTanimoto.csv')

# 5. Maximum Common Substructure (MCS)
# use atom counts to determine if a significant core is shared
mcsDf = pd.DataFrame(0, index=names, columns=names)
mcsThreshold = thresholdMCS

for i, rowName in enumerate(names):
    m1 = molSeries[rowName]
    if not m1: continue
    for j, colName in enumerate(names):
        if i >= j: continue # save compute by only comparing top right
        m2 = molSeries[colName]
        if m2:
            res = rdFMCS.FindMCS([m1, m2])
            # Similarity in this case is number of shared atoms / atoms in the smaller molecule
            overlap = res.numAtoms / min(m1.GetNumAtoms(), m2.GetNumAtoms())
            if overlap >= mcsThreshold:
                mcsDf.iloc[i, j] = 1
                mcsDf.iloc[j, i] = 1

save_matrix(mcsDf, 'mcsOverlap.csv')

```




```{r echo=FALSE, include=TRUE, message=FALSE, fig.cap="Cross reactivity matrix using direct matching algorithm"}
#| label: fig-huttenChart
#| fig-width: 10
#| fig-height: 10
plotAbxHeatmap("Original.csv", "Hutten et al.", colorRamp2(c(-1, 0, 1, 2, 3), c("#808080", "#A9D08F","#FF0000","#CF734F","#FFFFFF")))
```


# Direct Matching (PenicillinX original)

```{r DirectMatchChart, echo=FALSE, include=TRUE, message=FALSE, fig.cap="Cross reactivity matrix using direct matching algorithm"}
#| fig-width: 10
#| fig-height: 10
plotAbxHeatmap("directMatch.csv", "Direct Matching", colorRamp2(c(-1, 0, 1), c("#808080", "#A9D08F", "#FF0000")))
directMatchStats <- calculateStats("directMatch.csv", huttenMat) # 'mat' is from Hutten et al.
```


# Dice Similarity with Morgan Fingerprinting

```{r morganDiceChart, include=TRUE, echo=FALSE, fig.cap="Cross reactivity matrix using Morgan fingerprinting with Dice Similarity "}
#| fig-width: 10
#| fig-height: 10
plotAbxHeatmap("morganDice.csv", "Morgan-Dice Similarity", colorRamp2(c(-1, 0, 1), c("#808080", "#A9D08F", "#FF0000")))
morganDiceStats <- calculateStats("morganDice.csv", huttenMat)
```


# Tanimoto Similarity with Morgan Fingerprinting

```{r morganTanimotoChart, include=TRUE, echo=FALSE, fig.cap="Cross reactivity matrix using Morgan fingerprinting with Tanimoto Similarity "}
#| fig-width: 10
#| fig-height: 10
plotAbxHeatmap("morganTanimoto.csv", "Morgan-Tanimoto Similarity", colorRamp2(c(-1, 0, 1), c("#808080", "#A9D08F", "#FF0000")))
morganTanimotoStats <- calculateStats("morganTanimoto.csv", huttenMat)
```


# Pharmacocore with Morgan Fingerprinting

```{r PharmacoDice, include=TRUE, echo=FALSE, fig.cap="Cross reactivity matrix using Morgan fingerprinting with Pharmacocores"}
#| fig-width: 10
#| fig-height: 10
plotAbxHeatmap("pharmacoDice.csv", "Pharmacocore Dice Similarity", colorRamp2(c(-1, 0, 1), c("#808080", "#A9D08F", "#FF0000")))
pharmacoDiceStats <- calculateStats("pharmacoDice.csv", huttenMat)
```

# MACCS Tanimoto

```{r MACCSTanimoto, include=TRUE, echo=FALSE, fig.cap="Cross reactivity matrix using MACCS keys with Tanimoto Similarity "}
#| fig-width: 10
#| fig-height: 10
plotAbxHeatmap("maccsTanimoto.csv", "MACCS Tanimoto", colorRamp2(c(-1, 0, 1), c("#808080", "#A9D08F", "#FF0000")))
maccsTanimotoStats <- calculateStats("maccsTanimoto.csv", huttenMat)
```


# MCS Overlap

```{r MCS, include=TRUE, echo=FALSE, fig.cap="Cross reactivity matrix using Maximum Common Substructure"}
#| fig-width: 10
#| fig-height: 10
plotAbxHeatmap("mcsOverlap.csv", "Maximum Common Substructure Overlap", colorRamp2(c(-1, 0, 1), c("#808080", "#A9D08F", "#FF0000")))
mcsStats <- calculateStats("mcsOverlap.csv", huttenMat)
```

```{python ensembleMaker, echo=FALSE}
import pandas as pd
import numpy as np

# Add existing charts
files = ["directMatch.csv", "morganTanimoto.csv", "morganDice.csv", 
         "pharmacoDice.csv", "maccsTanimoto.csv"]
# , "mcsOverlap.csv" - Currently removed as sensitivity+specificity lower than MACCS

# Set index
dataframes = [pd.read_csv(f, index_col=0) for f in files]

#set true if ANY matrix has a 1
ensembleMask = (dataframes[0] == 1)
for df in dataframes[1:]:
    ensembleMask |= (df == 1)

# Convert Boolean to Integers
ensembleDf = ensembleMask.astype(int)

# Adds the -1s back down middle
np.fill_diagonal(ensembleDf.values, -1)

ensembleDf.to_csv("ensembleMatch.csv")


```

# All algorithms combined

```{r ensemble, include=TRUE, echo=FALSE, fig.cap="Cross reactivity matrix using Maximum Common Substructure"}
#| fig-width: 10
#| fig-height: 10
plotAbxHeatmap("ensembleMatch.csv", "all algos combined (MCS currently removed)", colorRamp2(c(-1, 0, 1), c("#808080", "#A9D08F", "#FF0000")))
ensembleStats <- calculateStats("ensembleMatch.csv", huttenMat)
```

# Hutten sine discrepant

```{r echo=FALSE, include=TRUE, message=FALSE, fig.cap="Hutten with discrepants set to nil"}
#| label: fig-huttensineDiscrepant
#| fig-width: 10
#| fig-height: 10
plotAbxHeatmap("Original.csv", "Hutten et al.", colorRamp2(c(-1, 0, 1, 2, 3), c("#808080", "#A9D08F","#FF0000","#A9D08F","#FFFFFF")))
```


\newpage

# Performance Table

```{python, echo=FALSE, include=TRUE}
#| label: performance-table
#| tbl-cap: "Compared Performance of Each Algorithm"

from IPython.display import Markdown
from tabulate import tabulate

# Reminder: access R lists using the 'r.' prefix

dm = r.directMatchStats
md = r.morganDiceStats
mt = r.morganTanimotoStats
pd = r.pharmacoDiceStats
mcs = r.mcsStats
mct = r.maccsTanimotoStats
ens = r.ensembleStats
# Table rows
table = [
    ["Set Threshold", 'N/A', thresholdMD, thresholdMT,  thresholdPhD, thresholdMCS,  thresholdMAC, 'N/A'],   
    ["Possible Hits", dm['possibleHits'], md['possibleHits'], mt['possibleHits'], pd['possibleHits'], mcs['possibleHits'], mct["possibleHits"],ens['possibleHits']],
    ["Total Hits",    dm['totalHits'],    md['totalHits'],    mt['totalHits'],    pd['totalHits'], mcs['totalHits'], mct["totalHits"],ens['totalHits']],
    ["True +ve",dm['truePositives'],md['truePositives'],mt['truePositives'], pd['truePositives'], mcs['truePositives'], mct["truePositives"],ens['truePositives']],
    ["False +ve",dm['falsePositives'],md['falsePositives'],mt['falsePositives'],pd['falsePositives'], mcs['falsePositives'], mct["falsePositives"],ens['falsePositives']],
    ["False -ve",dm['falseNegatives'],md['falseNegatives'],mt['falseNegatives'],pd['falseNegatives'], mcs['falseNegatives'], mct["falseNegatives"],ens['falseNegatives']],
    ["no Data", dm['noDataHit'], md['noDataHit'], mt['noDataHit'], pd['noDataHit'], mcs['noDataHit'], mct["noDataHit"],ens['noDataHit']],
    ["Sensitivity", dm['sensitivity'], md['sensitivity'], mt['sensitivity'], pd['sensitivity'], mcs['sensitivity'], mct['sensitivity'],ens['sensitivity']],
    ["Specificity", dm['specificity'], md['specificity'], mt['specificity'], pd['specificity'], mcs['specificity'], mct['specificity'],ens['specificity']]
    ]



# headers
headers = ["Metric", "Direct Match", "Morgan Dice", "Morgan Tanimoto", "Pharmaco Dice", "MCS", "MACCS Tanimoto", "Ensemble"]

# Render the table
Markdown(tabulate(table, headers=headers, tablefmt="pipe"))
```

```{python, echo=FALSE, Include=TRUE}
import pandas as pd
import numpy as np

# 1. Load the data
# Use the same paths you used in your previous blocks
ensembleDf = pd.read_csv("ensembleMatch.csv", index_col=0)
originalDf = pd.read_csv("Original.csv", index_col=0)

# 2. Align the dataframes to ensure drug names match exactly across rows/cols
# This prevents errors if the drug order differs between files
common_idx = originalDf.index.intersection(ensembleDf.index)
common_cols = originalDf.columns.intersection(ensembleDf.columns)

orig_aligned = originalDf.loc[common_idx, common_cols]
ens_aligned = ensembleDf.loc[common_idx, common_cols]

# 3. Find where Ensemble=1 and Original=3
# This creates a boolean mask
noDataMask = (ens_aligned == 1) & (orig_aligned == 3)

# 4. Extract the names of the pairs
rows, cols = np.where(noDataMask)
noDataPairs = set()

for r, c in zip(rows, cols):
    # Sort to ensure (DrugA, DrugB) is the same as (DrugB, DrugA)
    pair = tuple(sorted([ens_aligned.index[r], ens_aligned.columns[c]]))
    noDataPairs.add(pair)

# 5. Output the results as text
print(f"Found {len(noDataPairs)} potential cross-reactions with no clinical data in Hutten et al.:\n")
for abx1, abx2 in sorted(list(noDataPairs)):
    print(f"- {abx1} and {abx2}")
```

