---
title: "PenicillinX-Project"
format:
  pdf:
    cite-method: citeproc
    link-citations: true
bibliography: My Library.bib
csl: https://www.zotero.org/styles/american-medical-association
toc: true
fig-dpi: 300
fig-pos: H
header-includes:
  - \usepackage[margin=1in]{geometry}
---

\newpage

# Hutten et al.

```{r echo=FALSE, include=TRUE, message=FALSE, fig.cap="Cross reactivity matrix re-created from Hutten et al."}
#| label: OriginalChart
library(ComplexHeatmap)
library(circlize)
# Make sure Gu are cited for the above libraries if published
# [@gu2014]
# [@gu2022]

# Legend for Original.csv/Hutten et al. | NB: These are unrelated to the numbers displayed in the corresponding plot from the original publication
#-1 = Same Antibiotic on X/Y | Grey
# 0 = No cross reaction | Green
# 1 = Cross reaction | Red
# 2 = Discrepancy between studies | Brown
# 3 = No data | White

huttenMat <- as.matrix(read.csv("Original.csv", row.names = 1, check.names = FALSE))
# Function to calculate the stats
calculateStats <- function(csvPath, originalMat) {
  algoMat <- as.matrix(read.csv(csvPath, row.names = 1, check.names = FALSE))
  algoMat <- algoMat[rownames(originalMat), colnames(originalMat)]
  
  list(
    possibleHits   = sum(originalMat == 1) /2,
    totalHits      = sum(algoMat == 1) /2,
    truePositives  = sum(originalMat == 1 & algoMat == 1) /2,
    trueNegatives  = sum(originalMat == 0 & algoMat == 0) /2,
    falsePositives = sum(originalMat == 0 & algoMat == 1) /2,
    falseNegatives = sum(originalMat == 1 & algoMat == 0) /2,
    noDataHit      = sum(originalMat == 3 & algoMat == 1) /2,
#   hitRate        = round((sum(originalMat == 1 & algoMat == 1) / sum(originalMat == 1)),  2),
    sensitivity    = round(sum(originalMat == 1 & algoMat == 1) / (sum(originalMat == 1 & algoMat == 1) + sum(originalMat == 1 & algoMat == 0)),3),
    specificity    = round(sum(originalMat == 0 & algoMat == 0) / (sum(originalMat == 0 & algoMat == 0) + sum(originalMat == 0 & algoMat == 1)),3),
    youdensJ       = round((sum(originalMat == 1 & algoMat == 1) / (sum(originalMat == 1 & algoMat == 1) + sum(originalMat == 1 & algoMat == 0))) + 
                   (sum(originalMat == 0 & algoMat == 0) / (sum(originalMat == 0 & algoMat == 0) + sum(originalMat == 0 & algoMat == 1))) - 1, 3)
  )
}


# Function to generate Heatmaps
plotAbxHeatmap <- function(csvPath, chartTitle, colMapping) {
  localMat <- as.matrix(read.csv(csvPath, row.names = 1, check.names = FALSE))
  
  Heatmap(localMat, 
          name = "Match", 
          col = colMapping,
          column_title = chartTitle,
          row_names_side = "left", 
          column_names_side = "top",
          cluster_rows = FALSE, 
          cluster_columns = FALSE,
          rect_gp = gpar(col = "black", lwd = 0.4),
          column_names_rot = 90, 
          row_names_gp = gpar(fontsize = 8),
          column_names_gp = gpar(fontsize = 8), 
          show_heatmap_legend = FALSE,
          
          layer_fun = function(j, i, x, y, width, height, fill) {
            nRow = nrow(localMat)
            nCol = ncol(localMat)
            vIdx = seq(5, nCol - 1, by = 5)
            if(length(vIdx) > 0) {
              grid.segments(x0 = unit(vIdx / nCol, "npc"), y0 = unit(0, "npc"),
                            x1 = unit(vIdx / nCol, "npc"), y1 = unit(1, "npc"),
                            gp = gpar(lwd = 1.4, col = "black"))
            }
            hIdx = seq(5, nRow - 1, by = 5)
            if(length(hIdx) > 0) {
              grid.segments(x0 = unit(0, "npc"), y0 = unit(1 - (hIdx / nRow), "npc"),
                            x1 = unit(1, "npc"), y1 = unit(1 - (hIdx / nRow), "npc"),
                            gp = gpar(lwd = 1.4, col = "black"))
            }
          })
}

plotAbxHeatmapWithNumbers <- function(colorCsvPath, numbersCsvPath, chartTitle, colMapping) {
  colorMat <- as.matrix(read.csv(colorCsvPath, row.names = 1, check.names = FALSE))
  numberMat <- as.matrix(read.csv(numbersCsvPath, row.names = 1, check.names = FALSE))
  
  Heatmap(colorMat, 
          name = "Match", 
          col = colMapping,
          column_title = chartTitle,
          row_names_side = "left", 
          column_names_side = "top",
          cluster_rows = FALSE, 
          cluster_columns = FALSE,
          rect_gp = gpar(col = "black", lwd = 0.4),
          column_names_rot = 90, 
          row_names_gp = gpar(fontsize = 8),
          column_names_gp = gpar(fontsize = 8), 
          show_heatmap_legend = FALSE,
          
          # This overlays the text from the numbers CSV
          cell_fun = function(j, i, x, y, width, height, fill) {
            if (numberMat[i, j] != -1) {
              grid.text(sprintf("%s", numberMat[i, j]), x, y, 
                        gp = gpar(fontsize = 7, col = "black"))
            }
          },
          
          layer_fun = function(j, i, x, y, width, height, fill) {
            nRow = nrow(colorMat)
            nCol = ncol(colorMat)
            vIdx = seq(5, nCol - 1, by = 5)
            if(length(vIdx) > 0) {
              grid.segments(x0 = unit(vIdx / nCol, "npc"), y0 = unit(0, "npc"),
                            x1 = unit(vIdx / nCol, "npc"), y1 = unit(1, "npc"),
                            gp = gpar(lwd = 1.4, col = "black"))
            }
            hIdx = seq(5, nRow - 1, by = 5)
            if(length(hIdx) > 0) {
              grid.segments(x0 = unit(0, "npc"), y0 = unit(1 - (hIdx / nRow), "npc"),
                            x1 = unit(1, "npc"), y1 = unit(1 - (hIdx / nRow), "npc"),
                            gp = gpar(lwd = 1.4, col = "black"))
            }
          })
}

```

@fig-huttenChart is a reconstruction of the matrix detailed in Hutten et al[@hutten2025] Below them are the various molecular matching algorithms and their respective matrices, the performance of each algorithm, as well as their combined performance matching against Hutten et al is displayed in a table at the end of the document. Matrices generated using R library complexheatmap[@gu2022;@gu2014]. *This matrix was reconstructed by hand, and though it was checked for consistency against the original, it can't be guaranteed that it is free from transcription error.*




```{python}
#| label: algoblock
#| echo: false
#| include: true
import pandas as pd
import numpy as np
from rdkit import Chem, DataStructs
from rdkit.Chem import rdFingerprintGenerator, MACCSkeys, rdFMCS
from rdkit.Chem.Pharm2D import Gobbi_Pharm2D, Generate
from rdkit.Chem import rdFingerprintGenerator

###### Adjust thresholds here #######################
thresholdMD = 0.4867 # 100% Specificity at 0.667 # 0.55 gets optimal J
thresholdMT = 0.56 # 100% Specificity at 0.667 # 0.56 optimal J
thresholdPhD = 0.67 # 100% Specificity at 0.667 # 0.67 gets optimal J
thresholdMAC = 0.7041 # 99.97% Specificity at 0.667 with Tani, Dice at 0.78 with greater sensitivity # 0.68 optimal J (0.807)
thresholdMCS = 0.8914 #  # 100% Specificity at 0.92 # 0.89 optimal J
thresholdAPD = 0.4709 # >.53 gets worse .47 optimal J with Distance 7
####### -------------------- ########################


# Alternative thresholds
##Ultimate random. Sens .882 spec .976. J .858, ND 20
#thresholdMD = 0.542 
#thresholdMT = 0.56  
#thresholdPhD = 0.829 
#thresholdMAC = 0.743
#thresholdMCS = 0.898
#thresholdAPD = 0.505

##Penultimate random. Sens .882 spec .974. J .857. ND 22
#thresholdMD = 0.523 
#thresholdMT = 0.56  
#thresholdPhD = 0.801 
#thresholdMAC = 0.75
#thresholdMCS = 0.893
#thresholdAPD = 0.484


# Pull original PenX antibiotic data

abxData = pd.read_csv("antibiotics_data.csv")
names = abxData["Antibiotic"].tolist()
molSeries = abxData.set_index("Antibiotic")["SMILESR1"].apply(
    lambda x: Chem.MolFromSmiles(x) if pd.notna(x) else None
)

def save_matrix(df, fileName):
    np.fill_diagonal(df.values, -1)
    df.to_csv(fileName, index_label='Antibiotic')

# Direct Match (ie the PenicillinX original)
directMatchDf = pd.DataFrame(0, index=names, columns=names)
for row in names:
    m1 = molSeries[row]
    if m1:
        directMatchDf.loc[row] = molSeries.apply(
            lambda m2: 1 if (m2 and (m1.HasSubstructMatch(m2) or m2.HasSubstructMatch(m1))) else 0
        )
save_matrix(directMatchDf, 'directMatch.csv')

# Morgan Fingerprints (both Dice & Tanimoto) 
fpGen = rdFingerprintGenerator.GetMorganGenerator(radius=3, fpSize=2048) # Changing radius to 1 lowers youden's J, 3 reduced one false positive.
fpSeries = molSeries.apply(lambda m: fpGen.GetCountFingerprint(m) if m else None)
# No longer using Tanimoto, clean this up at some point
for mode, threshold, fileName in [('Dice', thresholdMD, 'morganDice.csv'), ('Tanimoto', thresholdMT, 'morganTanimoto.csv')]:
    df = pd.DataFrame(0, index=names, columns=names)
    simFunc = DataStructs.DiceSimilarity if mode == 'Dice' else DataStructs.TanimotoSimilarity
    for row in names:
        f1 = fpSeries[row]
        if f1:
            df.loc[row] = fpSeries.apply(lambda f2: 1 if (f2 and simFunc(f1, f2) >= threshold) else 0)
    save_matrix(df, fileName)

# Pharmacophore Dice
pharmFpSeries = molSeries.apply(lambda m: Generate.Gen2DFingerprint(m, Gobbi_Pharm2D.factory) if m else None)
pharmDf = pd.DataFrame(0, index=names, columns=names)
for row in names:
    f1 = pharmFpSeries[row]
    if f1:
        pharmDf.loc[row] = pharmFpSeries.apply(
            lambda f2: 1 if (f2 and DataStructs.DiceSimilarity(f1, f2) >= thresholdPhD) else 0
        )
save_matrix(pharmDf, 'pharmacoDice.csv')

# MACCS Keys # remember to change all references to DICE at some point
maccsFpSeries = molSeries.apply(lambda m: MACCSkeys.GenMACCSKeys(m) if m else None)
maccsDf = pd.DataFrame(0, index=names, columns=names)
for row in names:
    f1 = maccsFpSeries[row]
    if f1:
        maccsDf.loc[row] = maccsFpSeries.apply(
            lambda f2: 1 if (f2 and DataStructs.DiceSimilarity(f1, f2) >= thresholdMAC) else 0
        )
save_matrix(maccsDf, 'maccsDice.csv')

# Maximum Common Substructure (MCS)
# use atom counts to check if a significant core is shared
mcsDf = pd.DataFrame(0, index=names, columns=names)
mcsThreshold = thresholdMCS
# todo: Test different switches as per https://www.rdkit.org/docs/source/rdkit.Chem.fmcs.fmcs.html
for i, rowName in enumerate(names):
    m1 = molSeries[rowName]
    if not m1: continue
    for j, colName in enumerate(names):
        if i >= j: continue # A trick to only check top right and mirror
        m2 = molSeries[colName]
        if m2:
            res = rdFMCS.FindMCS([m1, m2], completeRingsOnly=False, ringMatchesRingOnly=True) # tweak these parameters and check 
            # Similarity in this case is number of shared atoms / atoms in the smaller molecule
            overlap = res.numAtoms / min(m1.GetNumAtoms(), m2.GetNumAtoms())
            if overlap >= mcsThreshold:
                mcsDf.iloc[i, j] = 1
                mcsDf.iloc[j, i] = 1

save_matrix(mcsDf, 'mcsOverlap.csv')

# Atom pairs using Dice Similarity
apGen = rdFingerprintGenerator.GetAtomPairGenerator(minDistance=1, maxDistance=7) # D7 appears optimal
# Make the fingerprint
apSeries = molSeries.apply(lambda m: apGen.GetCountFingerprint(m) if m else None)

# Checking the similarity
apDf = pd.DataFrame(0, index=names, columns=names)
for row in names:
    f1 = apSeries[row]
    if f1:
        # Using Dice Similarity to compare the count fingerprints
        apDf.loc[row] = apSeries.apply(
            lambda f2: 1 if (f2 and DataStructs.DiceSimilarity(f1, f2) >= thresholdAPD) else 0
        )

save_matrix(apDf, 'atomPairsDice.csv')

```

```{python}
#| label: PermissiveMatching
#| echo: false
#| include: true
import pandas as pd
import numpy as np
import re
from rdkit import Chem

# Pull my data
abxData = pd.read_csv("antibiotics_data.csv")
names = abxData["Antibiotic"].tolist()
molSeries = abxData.set_index("Antibiotic")["SMILESR1"].apply(
    lambda x: Chem.MolFromSmiles(x) if pd.notna(x) else None
)

def get_clinical_permissive_smarts(mol):
    if mol is None: return None
    
    # Generate SMARTS from the mol, initially from the SMILES
    smarts = Chem.MolToSmarts(mol)
    
    #Halogen Permissiveness trick
    # [#9]=F, [#17]=Cl, [#35]=Br, [#53]=I
    smarts = re.sub(r"\[#(9|17|35|53)\]", "[F,Cl,Br,I]", smarts)
    
    # Linker Permissiveness trick see: pen G vs Pen VK
    # regex looks for ether bonds and treats them as equivalent to a methylene
    smarts = smarts.replace("-[#8]-", "~") # Need to check it doesn't interfere with ester bonds, or if it even matters
    
    return smarts

# Pre-generate the patterns
patternSeries = molSeries.apply(get_clinical_permissive_smarts)

# Initialise Symmetric Matrix
permissiveMatchDf = pd.DataFrame(0, index=names, columns=names)

# Run the Symmetric Matching Loop
for i, rowName in enumerate(names):
    rowMol = molSeries[rowName]
    rowQueryStr = patternSeries[rowName]
    if rowMol is None or rowQueryStr is None: continue
    
    for j in range(i + 1, len(names)):
        colName = names[j]
        colMol = molSeries[colName]
        colQueryStr = patternSeries[colName]
        
        if colMol is None or colQueryStr is None: continue
        
        # Load queries from the permissive clinical strings
        queryRow = Chem.MolFromSmarts(rowQueryStr)
        queryCol = Chem.MolFromSmarts(colQueryStr)
        
        # Symmetric check: A matches query for B OR B matches query for A
        if (queryCol and rowMol.HasSubstructMatch(queryCol)) or \
           (queryRow and colMol.HasSubstructMatch(queryRow)):
            permissiveMatchDf.iloc[i, j] = 1
            permissiveMatchDf.iloc[j, i] = 1

def save_matrix(df, fileName):
    np.fill_diagonal(df.values, -1)
    df.to_csv(fileName, index_label='Antibiotic')

save_matrix(permissiveMatchDf, 'permissiveDirectMatch.csv')
```



```{r echo=FALSE, include=TRUE, message=FALSE, fig.cap="Cross reactivity matrix manually re-created from Hutten et al."}
#| label: fig-huttenChart
#| fig-width: 10
#| fig-height: 10

huttenColors <- colorRamp2(c(-1, 0, 1, 2, 3), 
                           c("#808080", "#A9D08F", "#FF0000", "#CF734F", "#FFFFFF"))

plotAbxHeatmapWithNumbers(
  colorCsvPath = "Original.csv", # Hutten's Cross reactivity values
  numbersCsvPath = "OriginalNumbered.csv", # The numbers inside Hutten et al's chart
  chartTitle = "Hutten et al.", 
  colMapping = huttenColors
)


#plotAbxHeatmap("Original.csv", "Hutten et al.", colorRamp2(c(-1, 0, 1, 2, 3), c("#808080", "#A9D08F","#FF0000","#CF734F","#FFFFFF")))
```
\newpage

# Direct Matching (PenicillinX Original)

The existing direct matching algorithm from PenicillinX web app.[@liamjones2025] Compares molecular structures directly using graph based comparisons of side chain SMILES. Sensitivity is limited, and often does not account for R1 alterations in side chain moieties that do not change allergenic potential.

```{r DirectMatchChart, echo=FALSE, include=TRUE, message=FALSE, fig.cap="Cross reactivity matrix using PenicillinX direct matching algorithm"}
#| fig-width: 10
#| fig-height: 10
plotAbxHeatmap("directMatch.csv", "Direct Matching", colorRamp2(c(-1, 0, 1), c("#808080", "#A9D08F", "#FF0000")))
directMatchStats <- calculateStats("directMatch.csv", huttenMat) # 'mat' is from Hutten et al.

```


\newpage
# Permissive Direct Matching (PDM) using SMARTS substitutions

A Matching algorithm similar to original PenicillinX,[@liamjones2025] but instead using SMARTS pattern matching.[@daylightchemicalinformationsystemsinc11] It implements bioisosteric substitution for features with similar expected behaviour specifically treating halogens as a united class (i.e. flucloxacillin-dicloxacillin) as well as methylene and ether linkages. (i.e. Penicillin G-VK). Continued work on this method shows promise in increasing the sensitivity of the algorithm while minimising loss to specificity.

```{r PermissiveDirectMatchChart, echo=FALSE, include=TRUE, message=FALSE, fig.cap="Cross reactivity matrix using an improved, more pemissive matching algorithm"}
#| fig-width: 10
#| fig-height: 10
plotAbxHeatmap("permissiveDirectMatch.csv", "Permissive Direct Matching", colorRamp2(c(-1, 0, 1), c("#808080", "#A9D08F", "#FF0000")))
permissiveDirectMatchStats <- calculateStats("permissiveDirectMatch.csv", huttenMat) # 'mat' is from Hutten et al.
```

\newpage

# Dice Similarity with Morgan Fingerprinting

This method utilises Morgan circular fingerprints,[@morgan1965] which iteratively encodes the local atomic environment into a bit vector. These fingerprints are then compared between two side chains using the Dice Coefficient. The Dice coefficient assigns values to molecular structures, and weighs shared features more heaviliy than the total feature set, providing a measure of structural overlap. \begin{equation} S_{AB} = \frac{2C}{A + B} \end{equation} where S~AB~ is similarity coefficient, C is the common features between both molecules, and A & B are the features in each molecule.

```{r morganDiceChart, include=TRUE, echo=FALSE, fig.cap="Cross reactivity matrix using Morgan fingerprinting with Dice Similarity "}
#| fig-width: 10
#| fig-height: 10
plotAbxHeatmap("morganDice.csv", "Morgan-Dice Similarity", colorRamp2(c(-1, 0, 1), c("#808080", "#A9D08F", "#FF0000")))
morganDiceStats <- calculateStats("morganDice.csv", huttenMat)
```
\newpage

# Tanimoto Similarity with Morgan Fingerprinting

A variation of the previous algorithm that instead utilises the Tanimoto Coefficient. Similar in concept to the Dice Coefficient, but with a different weight applied to matching substructures. Previous modelling in this project indicated that the Dice Coefficient strictly dominates the Tanimoto Coefficient, in detecting cross reactivity. Thus analyses using Tanimoto have been excluded from the final analysis.
\begin{equation} S_{AB} = \frac{C}{A + B - C} \end{equation}

```{r morganTanimotoChart, include=TRUE, echo=FALSE, fig.cap="Cross reactivity matrix using Morgan fingerprinting with Tanimoto Similarity "}
#| fig-width: 10
#| fig-height: 10
plotAbxHeatmap("morganTanimoto.csv", "Morgan-Tanimoto Similarity", colorRamp2(c(-1, 0, 1), c("#808080", "#A9D08F", "#FF0000")))
morganTanimotoStats <- calculateStats("morganTanimoto.csv", huttenMat)
```


# Pharmacocore with Dice Similarity

Pharmacocore fingerprinting maps the chemical characteristics of atoms and functional groups, such as hydrogen bond acceptors/donors, aliphatics/aromtatics. The Dice coefficient is then applied to attempt to model how an IgE molecule may 'perceive' a side chain's binding potential regardless of non-functional differences in the molecular structure.[@cereto-massague2015]

```{r PharmacoDice, include=TRUE, echo=FALSE, fig.cap="Cross reactivity matrix using Morgan fingerprinting with Pharmacocores"}
#| fig-width: 10
#| fig-height: 10
plotAbxHeatmap("pharmacoDice.csv", "Pharmacocore Dice Similarity", colorRamp2(c(-1, 0, 1), c("#808080", "#A9D08F", "#FF0000")))
pharmacoDiceStats <- calculateStats("pharmacoDice.csv", huttenMat)
```

# MACCS with Dice Similarity
<!--Changed from Tanimoto, remember to change variable names.
-->

This method utilises Molecular Access System(MACCS) keys, a specific type of pre-defined, publicly available structural fragment definition. There are 166 public fragment definitions in the RDKit implementation. A bit string is generated with reference to the presence or absence of certain features, for example, < 3 oxygens, S-S bond, or ring of size 4, etc.[@greglandrum2023:@joseaniceto2023] This method attempts to capture broad structural classes and motifs rather than atom-by-atom circular fingerprints.

```{r MACCSDice, include=TRUE, echo=FALSE, fig.cap="Cross reactivity matrix using MACCS keys with Dice Similarity "}
#| fig-width: 10
#| fig-height: 10
plotAbxHeatmap("maccsDice.csv", "MACCS Dice Similarity", colorRamp2(c(-1, 0, 1), c("#808080", "#A9D08F", "#FF0000")))
maccsDiceStats <- calculateStats("maccsDice.csv", huttenMat)
```


# Maximum Common Substructure (MCS) Overlap

The MCS algorithm identifies the largest substructure shared by two side chains. It performs a graph-matching search to find the maximum number of identical atoms and bonds common to both molecules. A similarity score is calculated as the ratio of the shared atoms to the total atoms in the smaller of the two molecules.[@andrewdalke2012] Calculated as: \begin{equation}
S_{MCS} = \frac{|MCS(A,B)|}{\min(|A|, |B|)}
\end{equation}

```{r MCS, include=TRUE, echo=FALSE, fig.cap="Cross reactivity matrix using Maximum Common Substructure"}
#| fig-width: 10
#| fig-height: 10
plotAbxHeatmap("mcsOverlap.csv", "Maximum Common Substructure Overlap", colorRamp2(c(-1, 0, 1), c("#808080", "#A9D08F", "#FF0000")))
mcsStats <- calculateStats("mcsOverlap.csv", huttenMat)
```

# Atom Pairs with Dice Similarity

Add description later. Original paper[@carhart1985] Picard et al usage[@picard2019]

```{r atomPairsChart, include=TRUE, echo=FALSE, fig.cap="Cross reactivity matrix using Atom Pairs with Dice Similarity"}
#| fig-width: 10
#| fig-height: 10
plotAbxHeatmap("atomPairsDice.csv", "Atom Pairs Dice Similarity", colorRamp2(c(-1, 0, 1), c("#808080", "#A9D08F", "#FF0000")))
atomPairsStats <- calculateStats("atomPairsDice.csv", huttenMat)
```

```{python}
#| label: ensembleMaker #i.e. combo
#| echo: false
#| include: true
import pandas as pd
import numpy as np

# Add existing charts
files = ["directMatch.csv", "morganDice.csv", 
         "pharmacoDice.csv", "maccsDice.csv", "permissiveDirectMatch.csv", "atomPairsDice.csv","mcsOverlap.csv"]
# "mcsOverlap.csv" - Currently [readded for testing] as sensitivity+specificity lower than MACCS , "morganTanimoto.csv removed"

# Set index
dataframes = [pd.read_csv(f, index_col=0) for f in files]

#set true if ANY matrix has a 1
ensembleMask = (dataframes[0] == 1)
for df in dataframes[1:]:
    ensembleMask |= (df == 1)

# Convert Boolean to Integers
ensembleDf = ensembleMask.astype(int)

# Adds the -1s back down middle
np.fill_diagonal(ensembleDf.values, -1)

ensembleDf.to_csv("ensembleMatch.csv")


```

```{python}
#| label: ensembleMaker
#| echo: false
#| include: true
import pandas as pd
import numpy as np

# List of all algorithm CSVs
files = ["directMatch.csv", "morganDice.csv", "pharmacoDice.csv", 
         "maccsDice.csv", "permissiveDirectMatch.csv", 
         "atomPairsDice.csv", "mcsOverlap.csv"]

# Load all dataframes
dataframes = [pd.read_csv(f, index_col=0) for f in files]

# 1. Create the Binary Mask (Any match = 1) for the background colors
ensembleMask = (dataframes[0] == 1)
for df in dataframes[1:]:
    ensembleMask |= (df == 1)
ensembleDf = ensembleMask.astype(int)
np.fill_diagonal(ensembleDf.values, -1)
ensembleDf.to_csv("ensembleMatch.csv")

# 2. Create the Count Matrix (Sum of hits) for the labels
# We convert each DF to 0/1 and sum them up
ensembleCountDf = pd.DataFrame(0, index=ensembleDf.index, columns=ensembleDf.columns)
for df in dataframes:
    # Ensure we only count actual '1's (ignoring -1s)
    ensembleCountDf += (df == 1).astype(int)

# Set diagonal to -1 so the heatmap function knows to ignore/grey them out
np.fill_diagonal(ensembleCountDf.values, -1)
ensembleCountDf.to_csv("ensembleCounts.csv")
``


# All algorithms combined

A matrix generated with the combined hits of all previous algorithms, currently excluding Morgan Tanimoto due to poor performance (it does not find matches not already covered by other algorithms). MCS overlap has been re-added after further tweaking, having been removed in earlier iterations due to poor sensitivity. Atom-Pairs algorithm has been added 

```{r include=TRUE, echo=FALSE, fig.cap="Combined performance of all algorithms"}
#| label: fig-combo
#| fig-width: 10
#| fig-height: 10
plotAbxHeatmap("ensembleMatch.csv", "Combined Algorithm Performance", colorRamp2(c(-1, 0, 1), c("#808080", "#A9D08F", "#FF0000")))
ensembleStats <- calculateStats("ensembleMatch.csv", huttenMat)
```

# Hutten et al. without discrepants

A re-creation of the Hutten et al[@hutten2025] Matrix with missing values and discrepant matches changed from red and white to green respectively. Currently present purely to aid in visual comparisons.

```{r echo=FALSE, include=TRUE, message=FALSE, fig.cap="Hutten with discrepants and missing values set to green for better visual comparison"}
#| label: fig-huttensineDiscrepant
#| fig-width: 10
#| fig-height: 10
plotAbxHeatmap("Original.csv", "Hutten et al conformed.", colorRamp2(c(-1, 0, 1, 2, 3), c("#808080", "#A9D08F","#FF0000","#A9D08F","#A9D08F")))
```





\newpage

# Performance Table

A table detailing the performance of each algorithm with the final column demonstrating the performance of all algorithms combined. Youden's J is used as an overall performance metric to assist in gauging the cost/benefit of each change to an algorithm. Youden's J is calculated as:
\begin{equation}
J = \text{Sensitivity} + \text{Specificity} - 1
\end{equation}

*I expect there should be a way that I can automate the process of tweaking the thresholds for the similarity coefficients to achieve an optimum Youden's J. Thus far these tweaks have been done manually, then revised by observing the result.*


```{python}
#| label: tbl-performanceTable
#| echo: false
#| include: true
#| tbl-cap: "Compared performance of each algorithm Morgan Tanimoto not included in combined result. DM:Direct Matching, MD:Morgan-Dice, PDM:Permissive Direct Matching, PD:Pharmacocore-Dice, MCS:Maximum Common Substructure, APD:Atom Pairs-Dice."
#| tbl-pos: H

from IPython.display import Markdown
from tabulate import tabulate

# Reminder: access R lists using the 'r.' prefix

dm = r.directMatchStats
pdm = r.permissiveDirectMatchStats
md = r.morganDiceStats
mt = r.morganTanimotoStats
pd = r.pharmacoDiceStats
mcs = r.mcsStats
mct = r.maccsDiceStats
apd = r.atomPairsStats
ens = r.ensembleStats
# Table rows
table = [
    ["Threshold", 'N/A', thresholdMD, 'N/A', thresholdPhD, thresholdMCS, thresholdMAC, thresholdAPD, 'N/A'],   
    ["Possible Hits", dm['possibleHits'], md['possibleHits'], pdm['possibleHits'], pd['possibleHits'], mcs['possibleHits'], mct["possibleHits"], apd['possibleHits'], ens['possibleHits']],
    ["Total Hits", dm['totalHits'], md['totalHits'], pdm['totalHits'], pd['totalHits'], mcs['totalHits'], mct["totalHits"], apd['totalHits'], ens['totalHits']],
    ["True +ve", dm['truePositives'], md['truePositives'], pdm['truePositives'], pd['truePositives'], mcs['truePositives'], mct["truePositives"], apd["truePositives"], ens['truePositives']],
    ["False +ve", dm['falsePositives'], md['falsePositives'], pdm['falsePositives'], pd['falsePositives'], mcs['falsePositives'], mct["falsePositives"], apd["falsePositives"], ens['falsePositives']],
    ["False -ve", dm['falseNegatives'], md['falseNegatives'], pdm['falseNegatives'], pd['falseNegatives'], mcs['falseNegatives'], mct["falseNegatives"], apd["falseNegatives"], ens['falseNegatives']],
    ["no Data", dm['noDataHit'], md['noDataHit'], pdm['noDataHit'], pd['noDataHit'], mcs['noDataHit'], mct["noDataHit"], apd["noDataHit"], ens['noDataHit']],
    ["Sensitivity", dm['sensitivity'], md['sensitivity'], pdm['sensitivity'], pd['sensitivity'], mcs['sensitivity'], mct['sensitivity'], apd['sensitivity'], ens['sensitivity']],
    ["Specificity", dm['specificity'], md['specificity'], pdm['specificity'], pd['specificity'], mcs['specificity'], mct['specificity'], apd['specificity'], ens['specificity']]
]

headers = ["Metric", "DM", "MD", "PDM", "Ph-Dice", "MCS", "MACCS Dice","APD","Combo"]

# Creates the main table first
tableLatex = tabulate(table, headers=headers, tablefmt="latex_booktabs")

# This adds Youden's J row with a midrule separator
youdenrRow = f"{dm['youdensJ']} & {md['youdensJ']} & {pdm['youdensJ']} & {pd['youdensJ']} & {mcs['youdensJ']} & {mct['youdensJ']}& {apd['youdensJ']} & {ens['youdensJ']}"
tableLatex = tableLatex.replace(r"\bottomrule", 
                                   f"\\midrule\nYouden's J & {youdenrRow} \\\\\n\\bottomrule")

Markdown(tableLatex)

```




# Raw Results

A generated list of matches deemed to be false positives or negatives when compared to the Hutten et al. chart for auditing purposes. These can be plugged into PenicillinX[@liamjones2025] for visual inspection of the structures in order to generate hypotheses on why a particular algorithm may have incorrectly hit or missed potential interactions.

```{python}
#| label: pair-attribution-analysis
#| echo: false
#| output: asis

import pandas as pd
import numpy as np

# Loads the matrices
try:
    ensembleDf = pd.read_csv("ensembleMatch.csv", index_col=0)
    originalDf = pd.read_csv("Original.csv", index_col=0)
    
    # Define and load individual matrices
    algo_files = {
        "DM": "directMatch.csv",
        "PDM": "permissiveDirectMatch.csv",
        "Morgan": "morganDice.csv",
        "Pharmaco": "pharmacoDice.csv",
        "MCS": "mcsOverlap.csv",
        "MACCS": "maccsDice.csv",
        "AP":"atomPairsDice.csv"
    }

    algos = {}
    for name, file in algo_files.items():
        try:
            algos[name] = pd.read_csv(file, index_col=0)
        except FileNotFoundError:
            continue

    # Align DataFrames
    common_idx = originalDf.index.intersection(ensembleDf.index)
    common_cols = originalDf.columns.intersection(ensembleDf.columns)
    orig_aligned = originalDf.loc[common_idx, common_cols]
    ens_aligned = ensembleDf.loc[common_idx, common_cols]

    # Define helper inside the same scope
    def get_contributing_algos(row_name, col_name):
        """Finds which individual algorithms flagged 1 for this pair."""
        contributors = []
        for name, df in algos.items():
            if row_name in df.index and col_name in df.columns:
                if df.loc[row_name, col_name] == 1:
                    contributors.append(name)
        return ", ".join(contributors)

    def run_analysis(mask, title):
        rows, cols = np.where(mask)
        pairs = set()
        pair_details = {}

        for r, c in zip(rows, cols):
            r_name = ens_aligned.index[r]
            c_name = ens_aligned.columns[c]
            pair = tuple(sorted([r_name, c_name]))
            
            if pair not in pairs:
                pairs.add(pair)
                pair_details[pair] = get_contributing_algos(r_name, c_name)

        # LaTeX rendering for PDF
        print(r"\subsubsection{" + f"{len(pairs)} " + title + r"}")
        print(r"\begin{itemize}")
        for (abx1, abx2), contributors in sorted(pair_details.items()):
            contrib_str = f" \\textbf{{[{contributors}]}}" if contributors else " [None]"
            # Escape & for LaTeX
            abx_line = f"{abx1} and {abx2}".replace("&", r"\&")
            print(r"  \item " + f"{abx_line}{contrib_str}")
        print(r"\end{itemize}")
        print(r"\hrule")

    # Generate the Sections
    # Potential Cross-Reactions (No Data)
    nd_mask = (ens_aligned == 1) & (orig_aligned == 3)
    run_analysis(nd_mask, "Predicted cross-reactions with no clinical data in Hutten et al.")
    # False Positives
    fp_mask = (ens_aligned == 1) & (orig_aligned == 0)
    run_analysis(fp_mask, "Potential false positives as compared to Hutten et al.")

    # Discrepant positives
    dp_mask = (ens_aligned == 1) & (orig_aligned == 2)
    run_analysis(dp_mask, "Predicted cross-reactions where there is discrepancy according to Hutten et al." )

    # Discrepant Negatives
    dp_mask = (ens_aligned == 0) & (orig_aligned == 2)
    run_analysis(dp_mask, "Negative predictions where there is discrepancy according to Hutten et al." )

    # False Negatives
    fn_mask = (ens_aligned == 0) & (orig_aligned == 1)
    run_analysis(fn_mask, "Potential false negatives as compared to Hutten et al.")
    



except Exception as e:
    print(f"Error during analysis: {e}")
    
# True Positives Analysis
tp_mask = (ens_aligned == 1) & (orig_aligned == 1)

def run_tp_analysis(mask, title):
    rows, cols = np.where(mask)
    pairs = set()
    pair_details = {}

    for r, c in zip(rows, cols):
        r_name = ens_aligned.index[r]
        c_name = ens_aligned.columns[c]
        pair = tuple(sorted([r_name, c_name]))
        
        if pair not in pairs:
            # We want to see which algos contributed
            contributors = get_contributing_algos(r_name, c_name)
            # Focus on "Unique" hits or those including the new AP module
            pairs.add(pair)
            pair_details[pair] = contributors

    print(r"\subsubsection{" + f"{len(pairs)} " + title + r"}")
    print(r"\begin{itemize}")
    for (abx1, abx2), contributors in sorted(pair_details.items()):
        # Highlight pairs where AP was a factor
        is_ap_hit = "AP" in contributors
        style = r"\textbf" if is_ap_hit else r"\textit"
        
        contrib_str = f" {style}{{{'[' + contributors + ']'}}}"
        abx_line = f"{abx1} and {abx2}".replace("&", r"\&")
        print(r"  \item " + f"{abx_line}{contrib_str}")
    print(r"\end{itemize}")
    print(r"\hrule")

# Execute
run_tp_analysis(tp_mask, "True Positives in agreement with Hutten et al.")
    
```




```{python}
#| label: fig-combined-pr-curve
#| echo: false
#| output: asis

import pandas as pd
import numpy as np
import matplotlib.pyplot as plt
from rdkit import Chem, DataStructs
from rdkit.Chem import rdFingerprintGenerator, MACCSkeys, rdFMCS
from rdkit.Chem.Pharm2D import Gobbi_Pharm2D, Generate
from sklearn.metrics import precision_recall_curve

# Load Data
abxData = pd.read_csv("antibiotics_data.csv")
truth_df = pd.read_csv("Original.csv", index_col=0)
names = abxData["Antibiotic"].tolist()
molSeries = abxData.set_index("Antibiotic")["SMILESR1"].apply(
    lambda x: Chem.MolFromSmiles(x) if pd.notna(x) else None
)

# Setup Generators
fpGen = rdFingerprintGenerator.GetMorganGenerator(radius=3, fpSize=2048)
apGen = rdFingerprintGenerator.GetAtomPairGenerator(minDistance=1, maxDistance=7) 

# Calculate Raw Similarity Scores
y_true = []
scores = {
    "Morgan Dice": [],
    "Pharmaco Dice": [],
    "MACCS Dice": [],
    "Atom Pairs Dice": [],
    "MCS Overlap": []
}

# Pre-generate fingerprints for speed
morgan_fps = molSeries.apply(lambda m: fpGen.GetCountFingerprint(m) if m else None)
pharm_fps = molSeries.apply(lambda m: Generate.Gen2DFingerprint(m, Gobbi_Pharm2D.factory) if m else None)
maccs_fps = molSeries.apply(lambda m: MACCSkeys.GenMACCSKeys(m) if m else None)
ap_fps = molSeries.apply(lambda m: apGen.GetCountFingerprint(m) if m else None)

# Align with truth_df 
for i, name1 in enumerate(names):
    for j, name2 in enumerate(names):
        if i >= j: continue 
        
        truth = truth_df.loc[name1, name2]
        if truth not in [0, 1]: continue 
        
        y_true.append(truth)
        m1, m2 = molSeries[name1], molSeries[name2]
        
        # Calculate scores
        scores["Morgan Dice"].append(DataStructs.DiceSimilarity(morgan_fps[name1], morgan_fps[name2]) if (morgan_fps[name1] and morgan_fps[name2]) else 0)
        scores["Pharmaco Dice"].append(DataStructs.DiceSimilarity(pharm_fps[name1], pharm_fps[name2]) if (pharm_fps[name1] and pharm_fps[name2]) else 0)
        scores["MACCS Dice"].append(DataStructs.DiceSimilarity(maccs_fps[name1], maccs_fps[name2]) if (maccs_fps[name1] and maccs_fps[name2]) else 0)
        scores["Atom Pairs Dice"].append(DataStructs.DiceSimilarity(ap_fps[name1], ap_fps[name2]) if (ap_fps[name1] and ap_fps[name2]) else 0)
        
        # MCS Overlap
        if m1 and m2:
            res = rdFMCS.FindMCS([m1, m2], completeRingsOnly=False, ringMatchesRingOnly=True)
            overlap = res.numAtoms / min(m1.GetNumAtoms(), m2.GetNumAtoms())
            scores["MCS Overlap"].append(overlap)
        else:
            scores["MCS Overlap"].append(0)

# Plotting
plt.figure(figsize=(8, 6))
current_thresholds = {
    "Morgan Dice": thresholdMD,
    "Pharmaco Dice": thresholdPhD,
    "MACCS Dice": thresholdMAC,
    "Atom Pairs Dice": thresholdAPD,
    "MCS Overlap": thresholdMCS
}

for label, y_scores in scores.items():
    precision, recall, thresholds = precision_recall_curve(y_true, y_scores)
    line, = plt.plot(recall, precision, label=label, lw=2)
    color = line.get_color()
    
    # Mark the current manual threshold on the curve
    idx = np.argmin(np.abs(thresholds - current_thresholds[label]))
    x_point = recall[idx]
    y_point = precision[idx]
    plt.scatter(recall[idx], precision[idx], s=60, color=color, edgecolor='black', zorder=5)

# Labelling of dots
    plt.annotate(label, 
                 (x_point, y_point), 
                 textcoords="offset points", 
                 xytext=(5, 5), 
                 fontsize=9,
                 fontweight='bold',
                 color=color)

plt.xlabel('Recall (Sensitivity)')
plt.ylabel('Precision (1 - False Discovery Rate)')
plt.title('Combined Precision-Recall Curves for Thresholded Cross-Reactivity Algorithms')
plt.xlim(0.2, 1.0);
plt.ylim(0.3, 1.05);
plt.legend()
plt.grid(True, alpha=0.3)
plt.tight_layout()

# Save and print LaTeX
plt.savefig("combined_pr_curve.png", dpi=300)
plt.close() # CRITICAL: Prevents double rendering in Quarto

print(r"\begin{center}")
print(r"\includegraphics[width=0.9\textwidth]{combined_pr_curve.png}")
print(r"\end{center}")
```


<!--{python}
#| label: global-optimizer-differential evolution
#| echo: false
#| output: false

import numpy as np
from scipy.optimize import differential_evolution
import time

# Prepare data 
labels = list(scores.keys())
score_matrix = np.array([scores[l] for l in labels]).T
y_true_arr = np.array(y_true)

# Static PDM Baseline 
# Ensure pdm_preds_arr is available in your workspace
pos_count = np.sum(y_true_arr == 1)
neg_count = np.sum(y_true_arr == 0)

# Define the Objective Function for DE
def ensemble_objective(trial_thresholds):
    # The optimiser tries to min the return value
    # Trial_thresholds order: Morgan, Pharmaco, MACCS, AP, MCS
    thresh_vec = np.array(trial_thresholds)
    
    # Ensemble logic: PDM (static) OR thresholded scores
    preds = pdm_preds_arr | np.any(score_matrix >= thresh_vec, axis=1)
    
    tp = np.sum(preds & (y_true_arr == 1))
    fp = np.sum(preds & (y_true_arr == 0))
    
    sens = tp / pos_count
    spec = (neg_count - fp) / neg_count
    
    # to max J we min (1 - J)
    j_index = sens + spec - 1
    
    # anchor specificity with a penalty
    penalty = 0
    if spec < 0.98:
        penalty = (0.98 - spec) * 10 
        
    return (1 - j_index) + penalty

# Configure and Run the Optimizer
# Bounds: Morgan, Pharmaco, MACCS, AtomPairs, MCS
bounds = [(0.40, 0.80), (0.50, 0.95), (0.50, 0.95), (0.40, 0.80), (0.65, 0.98)]

start_time = time.time()
result = differential_evolution(
    ensemble_objective, 
    bounds, 
    strategy='best1bin', 
    popsize=20,     # Size of population
    mutation=(0.5, 1), 
    recombination=0.7,
    tol=0.01        # Finish when improvement is minimal
)

#Extract and Assign Winning Thresholds
win_t_vals = result.x
win_j = 1 - result.fun # Approximate if penalty is 0

# Final check of stats for the winner
preds = pdm_preds_arr | np.any(score_matrix >= win_t_vals, axis=1)
tp = np.sum(preds & (y_true_arr == 1))
fp = np.sum(preds & (y_true_arr == 0))
win_sens = tp / pos_count
win_spec = (neg_count - fp) / neg_count
win_j_final = win_sens + win_spec - 1

# Update the global variables for downstream blocks - deprecated remove
thresholdMD = win_t_vals[0]
thresholdPhD = win_t_vals[1]
thresholdMAC = win_t_vals[2]
thresholdAPD = win_t_vals[3]
thresholdMCS = win_t_vals[4]

# Output Summary
print(r"\begin{tcolorbox}[colback=green!5,colframe=green!75,title=Differential Evolution Global Optimization]")
print(fr"Optimizer converged in {time.time() - start_time:.1f} seconds.\\")
print(r"\begin{itemize}")
print(fr"  \item \textbf{{Optimized Ensemble J:}} {win_j_final:.4f}")
print(fr"  \item \textbf{{Sensitivity:}} {win_sens:.4f} | \textbf{{Specificity:}} {win_spec:.4f}")
print(r"\end{itemize}")
print(r"\hrulefill\\")
print(r"\textbf{Winning Thresholds:}")
print(r"\begin{description}")
for i, label in enumerate(labels):
    print(fr"  \item[{label}:] {win_t_vals[i]:.4f}")
print(r"\end{description}")
print(r"\end{tcolorbox}")
```
-->



<!-- Optimising algorithm, to switch on, add ``` in front of the bracketed python and remove this comment syntax. takes about 4 minutes to run 20 million iterations. Only run the block, do not try to render with quarto.
{python}
#| label: global-ensemble-optimizer
#| echo: false
#| output: asis

import numpy as np
import time

# Define the Global Optimizer Function
def run_global_search_vectorized(y_true, scores_dict, iterations=20000000):
    best_j = -1
    best_thresholds = {}
    
    labels = list(scores_dict.keys())
    # score_matrix shape: (n_pairs, 5)
    score_matrix = np.array([scores_dict[l] for l in labels]).T
    y_true_arr = np.array(y_true)
    
    # Pre-calculate constants for sensitivity/specificity to avoid repeated sums
    pos_count = np.sum(y_true_arr == 1)
    neg_count = np.sum(y_true_arr == 0)

    # Pre-generate ALL thresholds: 20M x 5 floats ~= 800MB RAM
    # Define bounds: Morgan, Pharmaco, MACCS, AtomPairs, MCS
    lows =  np.array([0.40, 0.50, 0.50, 0.40, 0.65])
    highs = np.array([0.75, 0.90, 0.90, 0.75, 0.95])
    all_thresholds = np.random.uniform(lows, highs, size=(iterations, len(labels)))

    print(f"Starting 20M iterations...")
    start_time = time.time()

    for i in range(iterations):
        thresh_vec = all_thresholds[i]
        
        # Vectorized comparison across the 5 algorithms for this specific drug pair set
        # 'preds' is a boolean array: True if ANY algo matches
        preds = np.any(score_matrix >= thresh_vec, axis=1)
        
        # Calculate TP and FP using fast logical masking
        tp = np.sum(preds & (y_true_arr == 1))
        fp = np.sum(preds & (y_true_arr == 0))
        
        sens = tp / pos_count
        spec = (neg_count - fp) / neg_count
        current_j = sens + spec - 1
        
        # Clinical constraint + High Score check
        if spec >= 0.96 and current_j > best_j:
            best_j = current_j
            best_thresholds = dict(zip(labels, thresh_vec))
            best_stats = (sens, spec)
            
        # Optional: Progress heartbeat every 5 million
        if (i + 1) % 5000000 == 0:
            elapsed = time.time() - start_time
            print(f"--- {i+1} iterations reached ({elapsed:.1f}s) ---")

    return best_thresholds, best_j, best_stats

#. Run the Optimization
win_t, win_j, win_stats = run_global_search_vectorized(y_true, scores)

# Store results for the rest of the document
thresholdMD = win_t["Morgan Dice"]
thresholdPhD = win_t["Pharmaco Dice"]
thresholdMAC = win_t["MACCS Dice"]
thresholdAPD = win_t["Atom Pairs Dice"]
thresholdMCS = win_t["MCS Overlap"]

# Output a summary
print(r"\begin{tcolorbox}[colback=blue!5,colframe=blue!75,title=Global Optimizer Results (20M Iterations)]")
print(fr"Completed in {time.time() - start_time:.1f} seconds.\\")
print(r"\begin{itemize}")
print(fr"  \item \textbf{{Ensemble Youden's J:}} {win_j:.4f}")
print(fr"  \item \textbf{{Sensitivity:}} {win_stats[0]:.4f} | \textbf{{Specificity:}} {win_stats[1]:.4f}")
print(r"\end{itemize}")
print(r"\hrulefill\\")
print(r"\textbf{Winning Thresholds:}")
print(r"\begin{description}")
for algo, val in win_t.items():
    print(fr"  \item[{algo}:] {val:.4f}")
print(r"\end{description}")
print(r"\end{tcolorbox}")
```

-->


\newpage 

# References





<!-- 
# Todo
- Standardise legacy Rmarkdown formatting to Quarto
- optimisation algorithm
- continue optimising permissive match
- read more recent papers


-->
