---
title: "PenicillinX-Project"
format: pdf
bibliography: My Library.bib
csl: https://www.zotero.org/styles/american-medical-association
toc: true
fig-dpi: 300
fig-pos: H
header-includes:
  - \usepackage[margin=1in]{geometry}
---

\newpage

# Hutten et al.

```{r echo=FALSE, include=TRUE, message=FALSE, fig.cap="Cross reactivity matrix re-created from Hutten et al."}
#| label: OriginalChart
library(ComplexHeatmap)
library(circlize)
# Make sure Gu are cited for the above libraries if published
# [@gu2014]
# [@gu2022]
huttenMat <- as.matrix(read.csv("~/PycharmProjects/PenicillinX-Project/Original.csv", row.names = 1, check.names = FALSE))
# Function to calculate the stats
calculateStats <- function(csvPath, originalMat) {
  algoMat <- as.matrix(read.csv(csvPath, row.names = 1, check.names = FALSE))
  algoMat <- algoMat[rownames(originalMat), colnames(originalMat)]
  
  list(
    possibleHits   = sum(originalMat == 1) /2,
    totalHits      = sum(algoMat == 1) /2,
    truePositives  = sum(originalMat == 1 & algoMat == 1) /2,
    trueNegatives  = sum(originalMat == 0 & algoMat == 0) /2,
    falsePositives = sum(originalMat == 0 & algoMat == 1) /2,
    falseNegatives = sum(originalMat == 1 & algoMat == 0) /2,
    noDataHit      = sum(originalMat == 3 & algoMat == 1) /2,
#   hitRate        = round((sum(originalMat == 1 & algoMat == 1) / sum(originalMat == 1)),  2),
    sensitivity    = round(sum(originalMat == 1 & algoMat == 1) / (sum(originalMat == 1 & algoMat == 1) + sum(originalMat == 1 & algoMat == 0)),2),
    specificity    = round(sum(originalMat == 0 & algoMat == 0) / (sum(originalMat == 0 & algoMat == 0) + sum(originalMat == 0 & algoMat == 1)),3)
  )
}


# Function to generate Heatmaps
plotAbxHeatmap <- function(csvPath, chartTitle, colMapping) {
  localMat <- as.matrix(read.csv(csvPath, row.names = 1, check.names = FALSE))
  
  Heatmap(localMat, 
          name = "Match", 
          col = colMapping,
          column_title = chartTitle,
          row_names_side = "left", 
          column_names_side = "top",
          cluster_rows = FALSE, 
          cluster_columns = FALSE,
          rect_gp = gpar(col = "black", lwd = 0.4),
          column_names_rot = 90, 
          row_names_gp = gpar(fontsize = 8),
          column_names_gp = gpar(fontsize = 8), 
          show_heatmap_legend = FALSE,
          
          layer_fun = function(j, i, x, y, width, height, fill) {
            nRow = nrow(localMat)
            nCol = ncol(localMat)
            vIdx = seq(5, nCol - 1, by = 5)
            if(length(vIdx) > 0) {
              grid.segments(x0 = unit(vIdx / nCol, "npc"), y0 = unit(0, "npc"),
                            x1 = unit(vIdx / nCol, "npc"), y1 = unit(1, "npc"),
                            gp = gpar(lwd = 1.4, col = "black"))
            }
            hIdx = seq(5, nRow - 1, by = 5)
            if(length(hIdx) > 0) {
              grid.segments(x0 = unit(0, "npc"), y0 = unit(1 - (hIdx / nRow), "npc"),
                            x1 = unit(1, "npc"), y1 = unit(1 - (hIdx / nRow), "npc"),
                            gp = gpar(lwd = 1.4, col = "black"))
            }
          })
}
```

@fig-huttenChart is a reconstruction of the matrix detailed in Hutten et al. Below them are the various molecular matching algorithms and their respective matrices, the performance of each algorithm, as well as their combined performance matching against Hutten et al. is displayed in a table at the end of the document.




```{python}
#| label: algoblock
#| echo: false
#| include: true
import pandas as pd
import numpy as np
from rdkit import Chem, DataStructs
from rdkit.Chem import rdFingerprintGenerator, MACCSkeys, rdFMCS
from rdkit.Chem.Pharm2D import Gobbi_Pharm2D, Generate

###### Adjust thresholds here #######################
thresholdMD = 0.667 # 100% Specificity at 0.667
thresholdMT = 0.667 # 100% Specificity at 0.667
thresholdPhD = 0.667 # 100% Specificity at 0.667
thresholdMAC = 0.667 # 100% Specificity at 0.667
thresholdMCS = 0.85 #  # 100% Specificity at TBD
####### -------------------- ########################

# Pull original PenX antibiotic data
abxData = pd.read_csv("antibiotics_data.csv")
names = abxData["Antibiotic"].tolist()
molSeries = abxData.set_index("Antibiotic")["SMILESR1"].apply(
    lambda x: Chem.MolFromSmiles(x) if pd.notna(x) else None
)

def save_matrix(df, fileName):
    np.fill_diagonal(df.values, -1)
    df.to_csv(fileName, index_label='Antibiotic')

# Direct Match (ie the PenicillinX original)
directMatchDf = pd.DataFrame(0, index=names, columns=names)
for row in names:
    m1 = molSeries[row]
    if m1:
        directMatchDf.loc[row] = molSeries.apply(
            lambda m2: 1 if (m2 and (m1.HasSubstructMatch(m2) or m2.HasSubstructMatch(m1))) else 0
        )
save_matrix(directMatchDf, 'directMatch.csv')

# Morgan Fingerprints (both Dice & Tanimoto)
fpGen = rdFingerprintGenerator.GetMorganGenerator(radius=2, fpSize=2048)
fpSeries = molSeries.apply(lambda m: fpGen.GetCountFingerprint(m) if m else None)

for mode, threshold, fileName in [('Dice', thresholdMD, 'morganDice.csv'), ('Tanimoto', thresholdMT, 'morganTanimoto.csv')]:
    df = pd.DataFrame(0, index=names, columns=names)
    simFunc = DataStructs.DiceSimilarity if mode == 'Dice' else DataStructs.TanimotoSimilarity
    for row in names:
        f1 = fpSeries[row]
        if f1:
            df.loc[row] = fpSeries.apply(lambda f2: 1 if (f2 and simFunc(f1, f2) >= threshold) else 0)
    save_matrix(df, fileName)

# Pharmacophore Dice
pharmFpSeries = molSeries.apply(lambda m: Generate.Gen2DFingerprint(m, Gobbi_Pharm2D.factory) if m else None)
pharmDf = pd.DataFrame(0, index=names, columns=names)
for row in names:
    f1 = pharmFpSeries[row]
    if f1:
        pharmDf.loc[row] = pharmFpSeries.apply(
            lambda f2: 1 if (f2 and DataStructs.DiceSimilarity(f1, f2) >= thresholdPhD) else 0
        )
save_matrix(pharmDf, 'pharmacoDice.csv')

# MACCS Keys 
maccsFpSeries = molSeries.apply(lambda m: MACCSkeys.GenMACCSKeys(m) if m else None)
maccsDf = pd.DataFrame(0, index=names, columns=names)
for row in names:
    f1 = maccsFpSeries[row]
    if f1:
        maccsDf.loc[row] = maccsFpSeries.apply(
            lambda f2: 1 if (f2 and DataStructs.TanimotoSimilarity(f1, f2) >= thresholdMAC) else 0
        )
save_matrix(maccsDf, 'maccsTanimoto.csv')

# Maximum Common Substructure (MCS)
# use atom counts to check if a significant core is shared
mcsDf = pd.DataFrame(0, index=names, columns=names)
mcsThreshold = thresholdMCS

for i, rowName in enumerate(names):
    m1 = molSeries[rowName]
    if not m1: continue
    for j, colName in enumerate(names):
        if i >= j: continue # A trick to only check top right and mirror
        m2 = molSeries[colName]
        if m2:
            res = rdFMCS.FindMCS([m1, m2])
            # Similarity in this case is number of shared atoms / atoms in the smaller molecule
            overlap = res.numAtoms / min(m1.GetNumAtoms(), m2.GetNumAtoms())
            if overlap >= mcsThreshold:
                mcsDf.iloc[i, j] = 1
                mcsDf.iloc[j, i] = 1

save_matrix(mcsDf, 'mcsOverlap.csv')

```

```{python}
#| label: Permissive-Matching
#| echo: false
#| include: true
import pandas as pd
import numpy as np
import re
from rdkit import Chem

# Pull my data
abxData = pd.read_csv("antibiotics_data.csv")
names = abxData["Antibiotic"].tolist()
molSeries = abxData.set_index("Antibiotic")["SMILESR1"].apply(
    lambda x: Chem.MolFromSmiles(x) if pd.notna(x) else None
)

def get_clinical_permissive_smarts(mol):
    if mol is None: return None
    
    # Generate SMARTS from the mol, initially from the SMILES
    smarts = Chem.MolToSmarts(mol)
    
    #Halogen Permissiveness trick
    # [#9]=F, [#17]=Cl, [#35]=Br, [#53]=I
    smarts = re.sub(r"\[#(9|17|35|53)\]", "[F,Cl,Br,I]", smarts)
    
    # Linker Permissiveness trick see: pen G vs Pen VK
    # regex looks for ether bonds and treats them as equivalent to a methylene
    smarts = smarts.replace("-[#8]-", "~") # Need to check it doesn't interfere with ester bonds, or if it even matters
    
    return smarts

# Pre-generate the patterns
patternSeries = molSeries.apply(get_clinical_permissive_smarts)

# Initialise Symmetric Matrix
permissiveMatchDf = pd.DataFrame(0, index=names, columns=names)

# Run the Symmetric Matching Loop
for i, rowName in enumerate(names):
    rowMol = molSeries[rowName]
    rowQueryStr = patternSeries[rowName]
    if rowMol is None or rowQueryStr is None: continue
    
    for j in range(i + 1, len(names)):
        colName = names[j]
        colMol = molSeries[colName]
        colQueryStr = patternSeries[colName]
        
        if colMol is None or colQueryStr is None: continue
        
        # Load queries from the permissive clinical strings
        queryRow = Chem.MolFromSmarts(rowQueryStr)
        queryCol = Chem.MolFromSmarts(colQueryStr)
        
        # Symmetric check: A matches B_query OR B matches A_query
        if (queryCol and rowMol.HasSubstructMatch(queryCol)) or \
           (queryRow and colMol.HasSubstructMatch(queryRow)):
            permissiveMatchDf.iloc[i, j] = 1
            permissiveMatchDf.iloc[j, i] = 1

def save_matrix(df, fileName):
    np.fill_diagonal(df.values, -1)
    df.to_csv(fileName, index_label='Antibiotic')

save_matrix(permissiveMatchDf, 'permissiveDirectMatch.csv')
```



```{r echo=FALSE, include=TRUE, message=FALSE, fig.cap="Cross reactivity matrix using direct matching algorithm"}
#| label: fig-huttenChart
#| fig-width: 10
#| fig-height: 10
plotAbxHeatmap("Original.csv", "Hutten et al.", colorRamp2(c(-1, 0, 1, 2, 3), c("#808080", "#A9D08F","#FF0000","#CF734F","#FFFFFF")))
```
\newpage


# Permissive Direct Matching using SMARTS substitutions

Matching algorithm similar to original PenicillinX, but instead using SMARTS queries and additionally substituting like atoms such as halogens (ie flucloxacillin-dicloxacillin) as well as methylene and ethers.

```{r PermissiveDirectMatchChart, echo=FALSE, include=TRUE, message=FALSE, fig.cap="Cross reactivity matrix using an improved, more pemissive matching algorithm"}
#| fig-width: 10
#| fig-height: 10
plotAbxHeatmap("permissiveDirectMatch.csv", "Permissive Direct Matching", colorRamp2(c(-1, 0, 1), c("#808080", "#A9D08F", "#FF0000")))
permissiveDirectMatchStats <- calculateStats("permissiveDirectMatch.csv", huttenMat) # 'mat' is from Hutten et al.
```
\newpage

# Direct Matching (PenicillinX original)

The existing direct matching algorithm from PenicillinX web app. Compares molecular structures directly using SMILES

```{r DirectMatchChart, echo=FALSE, include=TRUE, message=FALSE, fig.cap="Cross reactivity matrix using direct matching algorithm"}
#| fig-width: 10
#| fig-height: 10
plotAbxHeatmap("directMatch.csv", "Direct Matching", colorRamp2(c(-1, 0, 1), c("#808080", "#A9D08F", "#FF0000")))
directMatchStats <- calculateStats("directMatch.csv", huttenMat) # 'mat' is from Hutten et al.

```
\newpage

# Dice Similarity with Morgan Fingerprinting

Morgan fingerprinting[@morgan1965], which assigns numeric identifiers to each atom in an iterative manner. These fingerprints are then compared between two side chains. Dice similarity assigns values to molecular structures, and weights them based on their frequency. \begin{equation} S_{AB} = \frac{2C}{A + B} \end{equation} where SAB is similarity coefficient, C is the common features between both molecules, and A & B are the features in each molecule

```{r morganDiceChart, include=TRUE, echo=FALSE, fig.cap="Cross reactivity matrix using Morgan fingerprinting with Dice Similarity "}
#| fig-width: 10
#| fig-height: 10
plotAbxHeatmap("morganDice.csv", "Morgan-Dice Similarity", colorRamp2(c(-1, 0, 1), c("#808080", "#A9D08F", "#FF0000")))
morganDiceStats <- calculateStats("morganDice.csv", huttenMat)
```
\newpage

# Tanimoto Similarity with Morgan Fingerprinting

```{r morganTanimotoChart, include=TRUE, echo=FALSE, fig.cap="Cross reactivity matrix using Morgan fingerprinting with Tanimoto Similarity "}
#| fig-width: 10
#| fig-height: 10
plotAbxHeatmap("morganTanimoto.csv", "Morgan-Tanimoto Similarity", colorRamp2(c(-1, 0, 1), c("#808080", "#A9D08F", "#FF0000")))
morganTanimotoStats <- calculateStats("morganTanimoto.csv", huttenMat)
```


# Pharmacocore with Morgan Fingerprinting

```{r PharmacoDice, include=TRUE, echo=FALSE, fig.cap="Cross reactivity matrix using Morgan fingerprinting with Pharmacocores"}
#| fig-width: 10
#| fig-height: 10
plotAbxHeatmap("pharmacoDice.csv", "Pharmacocore Dice Similarity", colorRamp2(c(-1, 0, 1), c("#808080", "#A9D08F", "#FF0000")))
pharmacoDiceStats <- calculateStats("pharmacoDice.csv", huttenMat)
```

# MACCS Tanimoto

```{r MACCSTanimoto, include=TRUE, echo=FALSE, fig.cap="Cross reactivity matrix using MACCS keys with Tanimoto Similarity "}
#| fig-width: 10
#| fig-height: 10
plotAbxHeatmap("maccsTanimoto.csv", "MACCS Tanimoto", colorRamp2(c(-1, 0, 1), c("#808080", "#A9D08F", "#FF0000")))
maccsTanimotoStats <- calculateStats("maccsTanimoto.csv", huttenMat)
```


# MCS Overlap

```{r MCS, include=TRUE, echo=FALSE, fig.cap="Cross reactivity matrix using Maximum Common Substructure"}
#| fig-width: 10
#| fig-height: 10
plotAbxHeatmap("mcsOverlap.csv", "Maximum Common Substructure Overlap", colorRamp2(c(-1, 0, 1), c("#808080", "#A9D08F", "#FF0000")))
mcsStats <- calculateStats("mcsOverlap.csv", huttenMat)
```

```{python}
#| label: ensembleMaker
#| echo: false
#| include: true
import pandas as pd
import numpy as np

# Add existing charts
files = ["directMatch.csv", "morganTanimoto.csv", "morganDice.csv", 
         "pharmacoDice.csv", "maccsTanimoto.csv", "permissiveDirectMatch.csv",]
# , "mcsOverlap.csv" - Currently removed as sensitivity+specificity lower than MACCS

# Set index
dataframes = [pd.read_csv(f, index_col=0) for f in files]

#set true if ANY matrix has a 1
ensembleMask = (dataframes[0] == 1)
for df in dataframes[1:]:
    ensembleMask |= (df == 1)

# Convert Boolean to Integers
ensembleDf = ensembleMask.astype(int)

# Adds the -1s back down middle
np.fill_diagonal(ensembleDf.values, -1)

ensembleDf.to_csv("ensembleMatch.csv")


```

# All algorithms combined

```{r include=TRUE, echo=FALSE, fig.cap="Cross reactivity matrix using Maximum Common Substructure"}
#| label: fig-ensemble
#| fig-width: 10
#| fig-height: 10
plotAbxHeatmap("ensembleMatch.csv", "all algos combined (MCS currently removed)", colorRamp2(c(-1, 0, 1), c("#808080", "#A9D08F", "#FF0000")))
ensembleStats <- calculateStats("ensembleMatch.csv", huttenMat)
```

# Hutten et al. without discrepants

```{r echo=FALSE, include=TRUE, message=FALSE, fig.cap="Hutten with discrepants set to nil for better visual clarity"}
#| label: fig-huttensineDiscrepant
#| fig-width: 10
#| fig-height: 10
plotAbxHeatmap("Original.csv", "Hutten et al.", colorRamp2(c(-1, 0, 1, 2, 3), c("#808080", "#A9D08F","#FF0000","#A9D08F","#FFFFFF")))
```


\newpage

# Performance Table

```{python}
#| label: tbl-performance-table
#| echo: false
#| include: true
#| tbl-cap: "Compared Performance of Each Algorithm"

from IPython.display import Markdown
from tabulate import tabulate

# Reminder: access R lists using the 'r.' prefix

dm = r.directMatchStats
pdm = r.permissiveDirectMatchStats
md = r.morganDiceStats
mt = r.morganTanimotoStats
pd = r.pharmacoDiceStats
mcs = r.mcsStats
mct = r.maccsTanimotoStats
ens = r.ensembleStats
# Table rows
table = [
    ["Set Threshold", 'N/A', thresholdMD, thresholdMT,  thresholdPhD, thresholdMCS,  thresholdMAC, 'N/A'],   
    ["Possible Hits", dm['possibleHits'], md['possibleHits'], pdm['possibleHits'], pd['possibleHits'], mcs['possibleHits'], mct["possibleHits"],ens['possibleHits']],
    ["Total Hits",    dm['totalHits'],    md['totalHits'],    pdm['totalHits'],    pd['totalHits'], mcs['totalHits'], mct["totalHits"],ens['totalHits']],
    ["True +ve",dm['truePositives'],md['truePositives'],pdm['truePositives'], pd['truePositives'], mcs['truePositives'], mct["truePositives"],ens['truePositives']],
    ["False +ve",dm['falsePositives'],md['falsePositives'],pdm['falsePositives'],pd['falsePositives'], mcs['falsePositives'], mct["falsePositives"],ens['falsePositives']],
    ["False -ve",dm['falseNegatives'],md['falseNegatives'],pdm['falseNegatives'],pd['falseNegatives'], mcs['falseNegatives'], mct["falseNegatives"],ens['falseNegatives']],
    ["no Data", dm['noDataHit'], md['noDataHit'], pdm['noDataHit'], pd['noDataHit'], mcs['noDataHit'], mct["noDataHit"],ens['noDataHit']],
    ["Sensitivity", dm['sensitivity'], md['sensitivity'], pdm['sensitivity'], pd['sensitivity'], mcs['sensitivity'], mct['sensitivity'],ens['sensitivity']],
    ["Specificity", dm['specificity'], md['specificity'], pdm['specificity'], pd['specificity'], mcs['specificity'], mct['specificity'],ens['specificity']]
    ]



# headers
headers = ["Metric", "Direct Match", "Morgan Dice", "PDM", "Pharmaco Dice", "MCS", "MACCS Tanimoto", "Ensemble"]

# Render the table
Markdown(tabulate(table, headers=headers, tablefmt="pipe"))
```

```{python, echo=FALSE, Include=TRUE}
import pandas as pd
import numpy as np

# 1. Load the data
# Use the same paths you used in your previous blocks
ensembleDf = pd.read_csv("ensembleMatch.csv", index_col=0)
originalDf = pd.read_csv("Original.csv", index_col=0)

# 2. Align the dataframes to ensure drug names match exactly across rows/cols
# This prevents errors if the drug order differs between files
common_idx = originalDf.index.intersection(ensembleDf.index)
common_cols = originalDf.columns.intersection(ensembleDf.columns)

orig_aligned = originalDf.loc[common_idx, common_cols]
ens_aligned = ensembleDf.loc[common_idx, common_cols]

# 3. Find where Ensemble=1 and Original=3
# This creates a boolean mask
noDataMask = (ens_aligned == 1) & (orig_aligned == 3)
# 4. Extract the names of the pairs
rows, cols = np.where(noDataMask)
noDataPairs = set()

for r, c in zip(rows, cols):
    # Sort to ensure (DrugA, DrugB) is the same as (DrugB, DrugA)
    pair = tuple(sorted([ens_aligned.index[r], ens_aligned.columns[c]]))
    noDataPairs.add(pair)

# 5. Output the results as text
print(f"Found {len(noDataPairs)} potential cross-reactions with no clinical data in Hutten et al.:\n")
for abx1, abx2 in sorted(list(noDataPairs)):
    print(f"- {abx1} and {abx2}")
    
# The same but for false positives
FPDataMask = (ens_aligned == 1) & (orig_aligned == 0)
rows, cols = np.where(FPDataMask)
FPDataPairs = set()

for r, c in zip(rows, cols):
    # Sort to ensure (DrugA, DrugB) is the same as (DrugB, DrugA)
    pair = tuple(sorted([ens_aligned.index[r], ens_aligned.columns[c]]))
    FPDataPairs.add(pair)

# 5. Output the results as text
print(f"\nFound {len(FPDataPairs)} potential False positives as compared to Hutten et al.:\n")
for abx1, abx2 in sorted(list(FPDataPairs)):
    print(f"- {abx1} and {abx2}")

 # The same but for false negatives
FNDataMask = (ens_aligned == 0) & (orig_aligned == 1)
rows, cols = np.where(FNDataMask)
FNDataPairs = set()

for r, c in zip(rows, cols):
    # Sort to ensure (DrugA, DrugB) is the same as (DrugB, DrugA)
    pair = tuple(sorted([ens_aligned.index[r], ens_aligned.columns[c]]))
    FNDataPairs.add(pair)

# 5. Output the results as text
print(f"\nFound {len(FNDataPairs)} potential False negatives as compared to Hutten et al.:\n")
for abx1, abx2 in sorted(list(FNDataPairs)):
    print(f"- {abx1} and {abx2}") 
```

